### 2014-04-15 paperFIE.tex
Primera versión del paper en Español

\documentclass[a4paper,10pt]{article}
%\usepackage[utf8x]{inputenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

%opening
\title{Detección automática de malos hábitos de programación con Scratch}
\author{Gregorio Robles, Jesús Moreno}

\begin{document}

\maketitle

\begin{abstract}
Este artículo describe un método de detección automática de dos malos hábitos de programación muy presentes en estudiantes que aprenden a programar con Scratch, como son la repetición de código y el nombrado de objetos, y plantea ideas para tratar de
evitar este tipo de situaciones.

\end{abstract}

\section{Introduction}
Scratch \cite{resnick2009scratch} es un entorno de programación que incluye un lenguaje de programación visual pensado para niños desde 6 años de edad, un entorno de desarrollo y un sitio web donde la comunidad puede alojar sus proyectos, ejecutar, estudiar y reutilizar otros programas, y compartir ideas o sugerencias. Aunque existen otros lenguajes de programación visuales similares, Scratch es sin duda el que más éxito ha tenido y el que se está usando de forma más habitual en todo el mundo, con más de tres millones de usuarios registrados en su web y más de cinco millones de proyectos compartidos en su repositorio \footnote{http://scratch.mit.edu/statistics/}.
\paragraph{}Scratch está siendo utilizado tanto en actividades extraescolares \cite{maloney2008programming, kafai2010entering} o campamentos de verano \cite{adams2010scratching, franklin2013assessment}, como en escuelas \cite{wilson2012evaluation}, institutos \cite{meerbaum2013learning} e incluso universidades \cite{wolz2009starting, malan2007scratch} de todo el mundo, y se ha documentado su capacidad no solo para acercar la programación y la informática a niños y jóvenes con el objetivo de que en el futuro se planteen continuar sus estudios en esta disciplina, sino también su éxito para enseñar conceptos de programación tanto básicos como avanzados.
\paragraph{}Sin embargo, también es posible encontrar trabajos que han detectado ciertos malos hábitos de programación en los estudiantes que aprenden a programar en este entorno \cite{meerbaum2011habits}. En esta línea, en nuestro trabajo como docentes hemos detectado otros malos hábitos o manías que, a pesar de insistir en ellas para tratar de evitarlas, vemos cómo se repiten constantemente en los alumonos que aprenden a programar en este entorno. Como estos hábitos son contrarios a las recomendaciones básicas de la programación, nuestro estudio trata de detectar si son comunes también en los proyectos compartidos en el repositorio y plantear ideas para tratar de evitar este tipo de situaciones.
\paragraph{}En la sección 2 se realiza un repaso sobre Scratch y se revisan trabajos que estudian los aprendizajes demostrados por estudiantes que aprenden a programar en este entorno, así como estudios que automatizan la evaluación automática de estos aprendizajes. En la sección 3 se describe la metodología seguida en nuestro trabajo. La sección 4 recoge los resultados obtenidos. Finalmente, en la sección 5  se explican las conclusiones de nuestro estudio y se ofrecen algunas ideas y sugerencias para futuros trabajos.

\section{Background}

\paragraph{}Scratch es un entorno de programación visual desarrollado por el grupo Lifelong Kindergarten del Media Laboratory del MIT que permite programar proyectos interactivos utilizando un enfoque drag-and-drop con diferentes piezas o bloques tipo \textit{Lego} \cite{resnick2009scratch}. Scratch se diseñó con el objetivo principal de animar a los jóvenes a programar, y la mayoría de sus usarios son aprendices noveles que no tienen experiencia previa en programación \cite{maloney2010scratch}. El entorno está diseñado para animar a los usuarios a que manipulen y modifiquen distintos tipos de elementos multimedia, de forma que niños y jóvenes pueden programar proyectos que les resultan interesantes y atractivos, como historias animadas, juegos y presentaciones interactivas \cite{maloney2008programming}
\paragraph{}Scratch también ofrece un sitio web en el que se publican los proyectos programados por los usuarios, se puede estudiar el código de cada programa, realizar comentarios y sugerencias y reutilizar el código de cualquier proyecto, por lo que cualquiera que quiera comenzar a programar tiene a su disposición una gran cantidad de recursos. Este enfoque, además, ofrece la oportunidad de aprender ciertas habilidades sociales relacionadas con la programación, como compartir y contribuir a la comunidad \cite{scaffidi2012skill}.
\paragraph{}Tal como se mencionaba en la introducción, es posible encontrar muchos trabajos \cite{maloney2008programming, franklin2013assessment,scaffidi2012skill,malan2007scratch} que demuestran que Scratch funciona como herramienta para aprender a programar, ya que existen casos de éxito en distintos nieveles educativos, entornos y disciplinas que muestran que los estudiantes aprenden conceptos como la interacción de usuario, sentencias condicionales, comunicación y sincronización, expresiones lógicas o representación de la información con variables.
\paragraph{}A pesar de estos buenos resultados, Meerbaum-Salant, Armoni y Ben-Ari \cite{meerbaum2011habits} ponen de manifiesto que los estudiantes que aprenden a programar en este entorno demuestran algunos malos hábitos que son contrarios a las buenas prácticas recomendadas. En concreto, demuestran una tendencia excesiva descomposición y un proceso de desarrollo totalmente de abajo a arriba que comienza con bloques individuales de Scratch.
\paragraph{}Con la idea de simplificar el proceso de evaluación del CT ayudando al evaluador con una herramienta de automatización de parte del proceso se creó Hairball \cite{boe2013hairball}, que es un analizador estático de proyectos Scratch inspirado en Lint\footnote{http://www.unix.com/man-page/FreeBSD/1/lint} que trata de detectar posibles errores de programación en un proyecto Scratch, como no inicializar el valor de una variable o el estado de un personaje, enviar mensajes que no son recibidos por ningún programa o incluir código que no llega a ejecutarse.

\section{Methodology}

\paragraph{}En nuestro trabajo como docentes con estudiantes de secundaria, hemos detectado que muchos estudiantes presentan dos malos hábitos de manera recurrente, a pesar de nuestra constante insistencia para tratar de evitarlos, que tienen que ver con el nombrado de objetos y con la repetición de código.

\paragraph{}Por una parte, es muy habitual comprobar que los estudiantes no cambian el nombre de los personajes, que el entorno bautiza como \textit{SpriteX}, donde X es un número incremental, de manera que al crecer el número de personajes de un proyecto resulta complicado saber a qué objeto se refiere una determinada sentencia, contribuyendo así a una mala legibilidad del código, ralentizando la programación del proyecto y complicando su depuración ante posibles errores.
\paragraph{}Por otro lado, es también muy habitual observar a los estudiantes repetir código en un mismo proyecto, a veces incluso en los programas de un mismo personaje. De esta forma, además de complicar el mantenimiento y la actualización del código, no se trabajan los conceptos de abstracción y modularización, dos componentes fundamentales del desarrollo del pensamiento computacional \cite{wing2008computational}.
\paragraph{}Con el objetivo de detectar estos dos comportamientos en los proyectos Scratch, se han desarrollado dos plug-ins para el entorno Hairball \cite{boe2013hairball}, que a su vez hace uso de la biblioteca implementada en python \textit{kurt}\footnote{https://github.com/blob8108/kurt}, para automatizar la localización de este tipo de prácticas en programas desarrollados en este entorno.
\paragraph{}El plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analiza un proyecto Scratch para comprobar si los nombre de los distintos personajes que aparecen en el proyecto comienzan con el string \textit{Sprite}, lo que indica que el usuario no ha modificado este nombre y ha dejado el que el entorno asigna por defecto a cada nuevo personaje añadido al proyecto.
\paragraph{}El plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analiza un proyecto Scratch para tratar de localizar programas completos repetidos dentro de un proyecto. Para ello recorre cada programa del proyecto y obtiene los \textit{tokens} de los bloques que lo forman, de manera que dos bloques que tan solo difieren en los valores que reciben son considerados iguales. Para ilustrar el funcionamiento de este plug-in, la imagen\ref{fig:CodeRepetition1} muestra dos scripts que contienen los mismos bloques con la única diferencia de los valores recibidos por parámetro. 
\begin{figure}
  \centering
    \includegraphics{img/CodeRepetition1.png}
  \caption{Two scripts repeating code}
  \label{fig:CodeRepetition1}
\end{figure}

\paragraph{}Estos scripts al ser analizados por el plug-in y traducidos a \textit{tokens} serían considerados iguales, ya que su código sería traducido al siguiente conjunto de bloques:
\begin{verbatim}
 'set rotation style %s'
 'repeat %s%s'
 'move %s steps'
 'next costume'
 'if on edge, bounce'
 'wait %s secs'
\end{verbatim}
\paragraph{} Así, la forma correcta de implementar este proyecto habría sido, tal como se muestra en la figura \ref{fig:CodeRepetition2}, definiendo un nuevo conjunto de bloques que recibe varios valores por parámetro y reutilizando el bloque creado en ambos programas, con las mejoras que esta solución implicaría en lo relativo al mantenimiento, actualización y depuración del código.

\begin{figure}
  \centering
    \includegraphics{img/CodeRepetition2.png}
  \caption{Defining own blocks to avoid code repetition}
  \label{fig:CodeRepetition2}
\end{figure}

\paragraph{}Para probar el funcionamiento de los plug-ins desarrollados y comprobar hasta qué punto estos comportamientos que habíamos detectado en nuestro trabajo se repiten también en los proyectos del repositorio de Scratch, procedimos a descargar 100 proyectos del mismo. Como la posibilidad de que los usuarios puedan definir sus propios bloques, con el uso del bloque \textit{def\_block}, solo se encuentra disponible desde el lanzamiento de Scratch 2.0, los proyectos descargados son todos posteriores a esta fecha.

\section{Findings}
\paragraph{}Los datos obtenidos tras analizar los 100 proyectos descargados confirman que las observaciones realizadas con nuestros estudiantes son también muy comunes en los proyectos de la comunidad. 
\paragraph{}En relación con la falta de costumbre de renombrar los personajes y no utilizar el nombre por defecto que asigna el entorno, el 79\% de los proyectos analizados contenían al menos un personaje sin renombrar del tipo \textit{SpriteX}.
\paragraph{}Como ejemplo para ilustrar los inconvenientes asociados a esta mala práctica se muestra en la imagen \ref{fig:SpriteNaming} un script que resulta difícilmente legible, ya que es necesario recorrer el área de personajes para localizar a qué personaje se refiere un bloque determinado, lo que dificulta la depuración del código ante eventuales problemas o mejoras e, incluso, ralentiza la propia programación por parte del estudiante, que debe memorizar o comprobar el número de objeto sobre el que quiere realizar una acción determinada.
\begin{figure}
  \centering
    \includegraphics{img/SpriteNaming.png}
  \caption{Sprites with default names}
  \label{fig:SpriteNaming}
\end{figure}
\paragraph{}En cuanto a la repetición de código, el 62\% de los proyectos analizados contenían al menos un programa repetido. De hecho, tan solo el 17\% de los proyectos analizados hacían uso del bloque \textit{def\_block} para definirse un bloque propio y poder reutilizarlo en otras partes del proyecto.

\section{Conclussions}

\paragraph{} En lo relativo al nombrado de objetos, es curioso comprobar que los usuarios de Scratch sí nombran correctamente, es decir, de forma significativa, las variables que utilizan en sus proyectos. En nuestra opinión esto se debe a que cuando se crea una variable es obligatorio asignarle un nombre, ya que el entorno no las nombra por defecto. Sin embargo, al crear un nuevo personaje se le asinga el nombre \textit{SpriteX} de manera automática. Quizás una modificación del entorno para que obligue al usuario a nombrar a los nuevos objetos eliminaría esta mala práctica.
\paragraph{} En relación a la repetición de código y el poco uso que los usuarios de Scratch hacen de la definición de bloques propios (el equivalente a la creación de métodos o procedimientos en otros lenguajes), quizás trabajos como el realizado por Seiter y Foreman \cite{seiter2013modeling} en el que se pone de manifiesto que la capacidad de abstracción y la modularización son conceptos y habilidadades que parecen desarrollarse a partir de cierta edad, puedan explicar los bajos índices de uso de esta funcionalidad, ya que una parte importante de la comunidad Scratch está formada por niños menores de 10 años\footnote{http://scratch.mit.edu/statistics/\#age}. Futuros trabajos deberían seguir investigando en esta línea para continuar avanzando en la definición de un marco de trabajo apropiado para distintas edades y para comprobar si los usuarios que aprenden en línea y aquellos que lo hacen en un entorno reglado demuestran índices distintos de uso de esta funcionalidad.

\newpage
\bibliography{scratch}
\bibliographystyle{unsrt}
\end{document}

### 2014-04-15 paperFIE_en.tex
Abstract and Introduction translated to English

--- 2014-04-15 paperFIE.tex
@@
 %opening
-\title{Detección automática de malos hábitos de programación con Scratch}
-\author{Gregorio Robles, Jesús Moreno}
+\title{Automatic detection of bad habits of programming in Scratch}
+\author{Jesús Moreno, Gregorio Robles}
 
@@
 \begin{abstract}
-Este artículo describe un método de detección automática de dos malos hábitos de programación muy presentes en estudiantes que aprenden a programar con Scratch, como son la repetición de código y el nombrado de objetos, y plantea ideas para tratar de
-evitar este tipo de situaciones.
+This paper describes a method for automatic detection of two very bad programming habits present in students learning to program with Scratch, such as the repetition of code and object naming, and raises ideas to try to avoid such situations.
 
@@
 \section{Introduction}
-Scratch \cite{resnick2009scratch} es un entorno de programación que incluye un lenguaje de programación visual pensado para niños desde 6 años de edad, un entorno de desarrollo y un sitio web donde la comunidad puede alojar sus proyectos, ejecutar, estudiar y reutilizar otros programas, y compartir ideas o sugerencias.
-Aunque existen otros lenguajes de programación visuales similares, Scratch es sin duda el que más éxito ha tenido y el que se está usando de forma más habitual en todo el mundo, con más de tres millones de usuarios registrados en su web y más de cinco millones de proyectos compartidos en su repositorio \footnote{http://scratch.mit.edu/statistics/}.
-\paragraph{}Scratch está siendo utilizado tanto en actividades extraescolares \cite{maloney2008programming, kafai2010entering} o campamentos de verano \cite{adams2010scratching, franklin2013assessment}, como en escuelas \cite{wilson2012evaluation}, institutos \cite{meerbaum2013learning} e incluso universidades \cite{wolz2009starting, malan2007scratch} de todo el mundo, y se ha documentado su capacidad no solo para acercar la programación y la informática a niños y jóvenes con el objetivo de que en el futuro se planteen continuar sus estudios en esta disciplina, sino también su éxito para enseñar conceptos de programación tanto básicos como avanzados.
-\paragraph{}Sin embargo, también es posible encontrar trabajos que han detectado ciertos malos hábitos de programación en los estudiantes que aprenden a programar en este entorno \cite{meerbaum2011habits}.
-En esta línea, en nuestro trabajo como docentes hemos detectado otros malos hábitos o manías que, a pesar de insistir en ellas para tratar de evitarlas, vemos cómo se repiten constantemente en los alumonos que aprenden a programar en este entorno.
-Como estos hábitos son contrarios a las recomendaciones básicas de la programación, nuestro estudio trata de detectar si son comunes también en los proyectos compartidos en el repositorio y plantear ideas para tratar de evitar este tipo de situaciones.
-\paragraph{}En la sección 2 se realiza un repaso sobre Scratch y se revisan trabajos que estudian los aprendizajes demostrados por estudiantes que aprenden a programar en este entorno, así como estudios que automatizan la evaluación automática de estos aprendizajes.
-En la sección 3 se describe la metodología seguida en nuestro trabajo.
-La sección 4 recoge los resultados obtenidos.
-Finalmente, en la sección 5  se explican las conclusiones de nuestro estudio y se ofrecen algunas ideas y sugerencias para futuros trabajos.
+Scratch \cite{resnick2009scratch}  is a programming environment that includes a visual programming language designed for children from 6 years old, a development environment and a website where the community can host their projects; run, study and reuse other programs; and share ideas or suggestions with other programmers.
+Although there are several similar visual programming languages​​, Scratch is undoubtedly the most successful and the one that is being used more regularly worldwide, with more than three million registered users on its website and more than five million shared project in its repository \footnote{http://scratch.mit.edu/statistics/}.
+\paragraph{}Scratch is being used both in extracurricular activities \cite{maloney2008programming, kafai2010entering} or summer camps \cite{adams2010scratching, franklin2013assessment}, as in schools \cite{wilson2012evaluation}, high schools \cite{meerbaum2013learning} and even universities worldwide \cite{wolz2009starting, malan2007scratch}.
+Its capacity to bring computer programming to children and youth with the aim that in the future they would be willing to continue their studies in this discipline, and also its success to teach both basic and advanced programming concepts has been documented.
+\paragraph{}However, it is also possible to find studies that have detected some bad programming habits in students learning to program in this environment \cite{meerbaum2011habits}.
+In this line, in our work as teachers we have detected other bad habits that, despite insisting our students to try to avoid them, we see how they are constantly repeated by pupils learning to program in Scratch.
+As these habits are contrary to the basic programming recommendations, our study tries to detect if they are also common in the projects shared on the Scratch repository and propose some ideas to try to avoid such situations.
+\paragraph{}In section 2 a brief overview of Scratch is presented, works that study the learning demonstrated by students learning to program in this environment are reviewed, and studies that automate the evaluation of this learning are exposed.
+In section 3, our work methodology is described.
+Section 4 presents the obtained results.
+Finally, in section 5 the conclusions of our study are discussed and some ideas and suggestions for future work are provided.
 

### 2014-04-15 paperFIE_en.tex
Fix errata

@@
 Scratch \cite{resnick2009scratch}  is a programming environment that includes a visual programming language designed for children from 6 years old, a development environment and a website where the community can host their projects; run, study and reuse other programs; and share ideas or suggestions with other programmers.
-Although there are several similar visual programming languages​​, Scratch is undoubtedly the most successful and the one that is being used more regularly worldwide, with more than three million registered users on its website and more than five million shared project in its repository \footnote{http://scratch.mit.edu/statistics/}.
+Although there are several similar visual programming languages, Scratch is undoubtedly the most successful and the one that is being used more regularly worldwide, with more than three million registered users on its website and more than five million shared project in its repository \footnote{http://scratch.mit.edu/statistics/}.
 \paragraph{}Scratch is being used both in extracurricular activities \cite{maloney2008programming, kafai2010entering} or summer camps \cite{adams2010scratching, franklin2013assessment}, as in schools \cite{wilson2012evaluation}, high schools \cite{meerbaum2013learning} and even universities worldwide \cite{wolz2009starting, malan2007scratch}.

### 2014-04-15 paperFIE_en.tex
Background translated to English

@@
 
-\paragraph{}Scratch es un entorno de programación visual desarrollado por el grupo Lifelong Kindergarten del Media Laboratory del MIT que permite programar proyectos interactivos utilizando un enfoque drag-and-drop con diferentes piezas o bloques tipo \textit{Lego} \cite{resnick2009scratch}.
-Scratch se diseñó con el objetivo principal de animar a los jóvenes a programar, y la mayoría de sus usarios son aprendices noveles que no tienen experiencia previa en programación \cite{maloney2010scratch}.
-El entorno está diseñado para animar a los usuarios a que manipulen y modifiquen distintos tipos de elementos multimedia, de forma que niños y jóvenes pueden programar proyectos que les resultan interesantes y atractivos, como historias animadas, juegos y presentaciones interactivas \cite{maloney2008programming}
-\paragraph{}Scratch también ofrece un sitio web en el que se publican los proyectos programados por los usuarios, se puede estudiar el código de cada programa, realizar comentarios y sugerencias y reutilizar el código de cualquier proyecto, por lo que cualquiera que quiera comenzar a programar tiene a su disposición una gran cantidad de recursos.
-Este enfoque, además, ofrece la oportunidad de aprender ciertas habilidades sociales relacionadas con la programación, como compartir y contribuir a la comunidad \cite{scaffidi2012skill}.
-\paragraph{}Tal como se mencionaba en la introducción, es posible encontrar muchos trabajos \cite{maloney2008programming, franklin2013assessment,scaffidi2012skill,malan2007scratch} que demuestran que Scratch funciona como herramienta para aprender a programar, ya que existen casos de éxito en distintos nieveles educativos, entornos y disciplinas que muestran que los estudiantes aprenden conceptos como la interacción de usuario, sentencias condicionales, comunicación y sincronización, expresiones lógicas o representación de la información con variables.
-\paragraph{}A pesar de estos buenos resultados, Meerbaum-Salant, Armoni y Ben-Ari \cite{meerbaum2011habits} ponen de manifiesto que los estudiantes que aprenden a programar en este entorno demuestran algunos malos hábitos que son contrarios a las buenas prácticas recomendadas.
-En concreto, demuestran una tendencia excesiva descomposición y un proceso de desarrollo totalmente de abajo a arriba que comienza con bloques individuales de Scratch.
-\paragraph{}Con la idea de simplificar el proceso de evaluación del CT ayudando al evaluador con una herramienta de automatización de parte del proceso se creó Hairball \cite{boe2013hairball}, que es un analizador estático de proyectos Scratch inspirado en Lint\footnote{http://www.unix.com/man-page/FreeBSD/1/lint} que trata de detectar posibles errores de programación en un proyecto Scratch, como no inicializar el valor de una variable o el estado de un personaje, enviar mensajes que no son recibidos por ningún programa o incluir código que no llega a ejecutarse.
+\paragraph{}Scratch is a visual programming environment developed by the Lifelong Kindergarten group at MIT Media Laboratory that allows to program interactive projects using a drag-and-drop approach with different pieces or \textit {Lego} style blocks \cite{resnick2009scratch}.
+Scratch was designed with the main goal of encouraging young people to program, and most of its users are novice learners with no prior programming experience \cite{maloney2010scratch}.
+The environment is designed to encourage users to manipulate and modify various media types, so that children and youth can create projects that they find interesting and attractive as animated stories, games or interactive presentations \cite{maloney2008programming}
+\paragraph{}Scratch also offers a website where the projects programmed by the community are published, the code of each program can be studied, and it is possible to make suggestions and reuse the code of any project, so there are lots of valuable resources available for anyone willing to start to program.
+This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community \cite{scaffidi2012skill}.
+\paragraph{}As mentioned in the introduction, many papers showing that Scratch works as a tool for learning to program can be found \cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, as there are success stories in different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
+\paragraph{}Despite these good results, Meerbaum-Salant, Armoni and Ben-Ari \cite{meerbaum2011habits} prove that students learning to program in this environment show some bad habits that are contrary to recommended programming practices.
+In particular, they show excessive decomposition tendency and a development process absolutly from bottom to top starting with individual blocks of Scratch.
+\paragraph{} Aiming to simplify the process of evaluating the Computational Thinking and help evaluators with an automation tool, Hairball was created \cite{boe2013hairball}.
+Hairball is a static analyzer of Scratch projects inspired by Lint \footnote{http:// www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program or including code that is never run.
 

### 2014-04-17 paperFIE_en.tex
Quick translation of the paper. Paragraphs starting with "--Sin retocar--" have not been "refined".

@@
 
-\paragraph{}En nuestro trabajo como docentes con estudiantes de secundaria, hemos detectado que muchos estudiantes presentan dos malos hábitos de manera recurrente, a pesar de nuestra constante insistencia para tratar de evitarlos, que tienen que ver con el nombrado de objetos y con la repetición de código.
+\paragraph{}--Sin retocar--In our work as teachers with high school students, we found that many students have two bad habits recurring basis, despite our constant insistence to try to avoid them having to do with the named object and code repetition.
 
-\paragraph{}Por una parte, es muy habitual comprobar que los estudiantes no cambian el nombre de los personajes, que el entorno bautiza como \textit{SpriteX}, donde X es un número incremental, de manera que al crecer el número de personajes de un proyecto resulta complicado saber a qué objeto se refiere una determinada sentencia, contribuyendo así a una mala legibilidad del código, ralentizando la programación del proyecto y complicando su depuración ante posibles errores.
-\paragraph{}Por otro lado, es también muy habitual observar a los estudiantes repetir código en un mismo proyecto, a veces incluso en los programas de un mismo personaje.
-De esta forma, además de complicar el mantenimiento y la actualización del código, no se trabajan los conceptos de abstracción y modularización, dos componentes fundamentales del desarrollo del pensamiento computacional \cite{wing2008computational}.
-\paragraph{}Con el objetivo de detectar estos dos comportamientos en los proyectos Scratch, se han desarrollado dos plug-ins para el entorno Hairball \cite{boe2013hairball}, que a su vez hace uso de la biblioteca implementada en python \textit{kurt}\footnote{https://github.com/blob8108/kurt}, para automatizar la localización de este tipo de prácticas en programas desarrollados en este entorno.
-\paragraph{}El plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analiza un proyecto Scratch para comprobar si los nombre de los distintos personajes que aparecen en el proyecto comienzan con el string \textit{Sprite}, lo que indica que el usuario no ha modificado este nombre y ha dejado el que el entorno asigna por defecto a cada nuevo personaje añadido al proyecto.
-\paragraph{}El plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analiza un proyecto Scratch para tratar de localizar programas completos repetidos dentro de un proyecto.
-Para ello recorre cada programa del proyecto y obtiene los \textit{tokens} de los bloques que lo forman, de manera que dos bloques que tan solo difieren en los valores que reciben son considerados iguales.
-Para ilustrar el funcionamiento de este plug-in, la imagen\ref{fig:CodeRepetition1} muestra dos scripts que contienen los mismos bloques con la única diferencia de los valores recibidos por parámetro.
+\paragraph{}--Sin retocar--On the one hand, it is very common to test students do not change the names of the characters, the environment christens \textit{SpriteX}, where X is an incremental number, so that with increasing number of characters in a project is difficult to know which object relates a given statement, contributing to poor code readability, slowing the project schedule and complicating debugging for possible errors.
+\paragraph{}--Sin retocar--On the other hand, is also very common to see students repeat code in the same project, sometimes even in the same character programs.
+Thus, further complicating the maintenance and updating of the code, not the concepts of abstraction and modularization will work, two key components of the development of computational thinking \cite{wing2008computational}.
+\paragraph{} --Sin retocar--In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment \cite{boe2013hairball}, which in turn makes use of the library implemented in python \textit{kurt}\footnote{https://github.com/blob8108/kurt}, to automate the positioning of this type of practice in programs developed in this environment.
+\paragraph{}--Sin retocar--The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check the names of the various characters in the project beginning with the string \ textit {Sprite}, indicating that the user has not changed its name and has left the system assigns a default every new character added to the project.
+\paragraph{}--Sin retocar--The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
+This program runs each project and gets the \textit{tokens} blocks that form, so that two blocks that differ only in the values received are considered equal.
+To illustrate the operation of this plug-in, the picture \ref{fig:CodeRepetition1} shows two scripts containing same blocks with the only difference being received by parameter values.
 \begin{figure}
@@
 
-\paragraph{}Estos scripts al ser analizados por el plug-in y traducidos a \textit{tokens} serían considerados iguales, ya que su código sería traducido al siguiente conjunto de bloques:
+\paragraph{}--Sin retocar--These scripts to be analyzed by the plug-in and translated into \textit{tokens} would be considered equal, because your code would be translated to the next set of blocks:
 \begin{verbatim}
@@
 \end{verbatim}
-\paragraph{} Así, la forma correcta de implementar este proyecto habría sido, tal como se muestra en la figura \ref{fig:CodeRepetition2}, definiendo un nuevo conjunto de bloques que recibe varios valores por parámetro y reutilizando el bloque creado en ambos programas, con las mejoras que esta solución implicaría en lo relativo al mantenimiento, actualización y depuración del código.
+\paragraph{} --Sin retocar-- Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks receiving multiple values per parameter and reusing the block created 	in both programs, with this solution would involve improvements as regards the maintenance, updating and debugging code.
 
@@
 
-\paragraph{}Para probar el funcionamiento de los plug-ins desarrollados y comprobar hasta qué punto estos comportamientos que habíamos detectado en nuestro trabajo se repiten también en los proyectos del repositorio de Scratch, procedimos a descargar 100 proyectos del mismo.
-Como la posibilidad de que los usuarios puedan definir sus propios bloques, con el uso del bloque \textit{def\_block}, solo se encuentra disponible desde el lanzamiento de Scratch 2.0, los proyectos descargados son todos posteriores a esta fecha.
+\paragraph{}--Sin retocar-- To test the functionality of the plug-ins developed and test how these behaviors that we detected in our work are also repeated in the projects repository Scratch, we proceeded to download 100 similar projects.
+As the possibility for users to define their own blocks, using block \textit{def\_block}, only available since the release of Scratch 2.0 downloaded projects are all after this date.
 
 \section{Findings}
-\paragraph{}Los datos obtenidos tras analizar los 100 proyectos descargados confirman que las observaciones realizadas con nuestros estudiantes son también muy comunes en los proyectos de la comunidad.
-\paragraph{}En relación con la falta de costumbre de renombrar los personajes y no utilizar el nombre por defecto que asigna el entorno, el 79\% de los proyectos analizados contenían al menos un personaje sin renombrar del tipo \textit{SpriteX}.
-\paragraph{}Como ejemplo para ilustrar los inconvenientes asociados a esta mala práctica se muestra en la imagen \ref{fig:SpriteNaming} un script que resulta difícilmente legible, ya que es necesario recorrer el área de personajes para localizar a qué personaje se refiere un bloque determinado, lo que dificulta la depuración del código ante eventuales problemas o mejoras e, incluso, ralentiza la propia programación por parte del estudiante, que debe memorizar o comprobar el número de objeto sobre el que quiere realizar una acción determinada.
+\paragraph{}--Sin retocar--
+The data obtained after analyzing 100 projects downloaded observations confirm that our students are also very common in community projects.
+\paragraph{}--Sin retocar--Regarding the lack of habit of renaming the characters and do not use the default name assigned by the environment, the 79 \% of the projects analyzed contained at least one character without renaming the type \textit{SpriteX}.
+\paragraph{}--Sin retocar--As an example to illustrate the drawbacks associated with this malpractice shown in picture \ref{fig:SpriteNaming}  a script that is difficult to read as it is necessary to explore the area to find characters which character refers to a particular block, making it difficult to debug the code before any problems or improvements and even slows own programming for the student, you should memorize or check the number of object you want to perform an action.
 \begin{figure}
@@
 \end{figure}
-\paragraph{}En cuanto a la repetición de código, el 62\% de los proyectos analizados contenían al menos un programa repetido.
-De hecho, tan solo el 17\% de los proyectos analizados hacían uso del bloque \textit{def\_block} para definirse un bloque propio y poder reutilizarlo en otras partes del proyecto.
+\paragraph{}--Sin retocar--As for the repetition code, 62\% of the projects analyzed contained at least one repeated program.
+In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define a block itself and can be reused in other parts of the project.
 
@@
 
-\paragraph{} En lo relativo al nombrado de objetos, es curioso comprobar que los usuarios de Scratch sí nombran correctamente, es decir, de forma significativa, las variables que utilizan en sus proyectos.
-En nuestra opinión esto se debe a que cuando se crea una variable es obligatorio asignarle un nombre, ya que el entorno no las nombra por defecto.
-Sin embargo, al crear un nuevo personaje se le asinga el nombre \textit{SpriteX} de manera automática.
-Quizás una modificación del entorno para que obligue al usuario a nombrar a los nuevos objetos eliminaría esta mala práctica.
-\paragraph{} En relación a la repetición de código y el poco uso que los usuarios de Scratch hacen de la definición de bloques propios (el equivalente a la creación de métodos o procedimientos en otros lenguajes), quizás trabajos como el realizado por Seiter y Foreman \cite{seiter2013modeling} en el que se pone de manifiesto que la capacidad de abstracción y la modularización son conceptos y habilidadades que parecen desarrollarse a partir de cierta edad, puedan explicar los bajos índices de uso de esta funcionalidad, ya que una parte importante de la comunidad Scratch está formada por niños menores de 10 años\footnote{http://scratch.mit.edu/statistics/\#age}.
-Futuros trabajos deberían seguir investigando en esta línea para continuar avanzando en la definición de un marco de trabajo apropiado para distintas edades y para comprobar si los usuarios que aprenden en línea y aquellos que lo hacen en un entorno reglado demuestran índices distintos de uso de esta funcionalidad.
+\paragraph{} --Sin retocar--Regarding the named object is curious that other users Scratch named correctly, ie, significantly, the variables used in their projects.
+In our opinion this is because when you create a variable is mandatory to assign a name, because the environment does not name by default.
+However, when creating a new character is asinga named \textit{SpriteX} automatically.
+Maybe a change of environment that requires the user to name the new objects eliminate this malpractice.
+\paragraph{} --Sin retocar-- Regarding the code repetition and low usage users Scratch make the definition of own blocks (the equivalent of creating methods or procedures in other languages), you may work as performed by Seiter and Foreman \cite {seiter2013modeling} in which shows the capacity for abstraction and modularization are concepts that seem habilidadades develop from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 \footnote{http://scratch.mit.edu/statistics/\#age}.
+Future studies should further investigate this line to continue advancing in the definition of an appropriate framework for different ages and to check whether users online learning and those who do in a regulated environment show rates than use this functionality .
 

### 2014-04-17 paperFIE_en.tex
Methodology section refined

@@
 
-\paragraph{}--Sin retocar--In our work as teachers with high school students, we found that many students have two bad habits recurring basis, despite our constant insistence to try to avoid them having to do with the named object and code repetition.
+\paragraph{}In our work as teachers with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits in a regular basis that have to do with named object and code repetition.
 
-\paragraph{}--Sin retocar--On the one hand, it is very common to test students do not change the names of the characters, the environment christens \textit{SpriteX}, where X is an incremental number, so that with increasing number of characters in a project is difficult to know which object relates a given statement, contributing to poor code readability, slowing the project schedule and complicating debugging for possible errors.
-\paragraph{}--Sin retocar--On the other hand, is also very common to see students repeat code in the same project, sometimes even in the same character programs.
-Thus, further complicating the maintenance and updating of the code, not the concepts of abstraction and modularization will work, two key components of the development of computational thinking \cite{wing2008computational}.
-\paragraph{} --Sin retocar--In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment \cite{boe2013hairball}, which in turn makes use of the library implemented in python \textit{kurt}\footnote{https://github.com/blob8108/kurt}, to automate the positioning of this type of practice in programs developed in this environment.
-\paragraph{}--Sin retocar--The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check the names of the various characters in the project beginning with the string \ textit {Sprite}, indicating that the user has not changed its name and has left the system assigns a default every new character added to the project.
-\paragraph{}--Sin retocar--The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
-This program runs each project and gets the \textit{tokens} blocks that form, so that two blocks that differ only in the values received are considered equal.
-To illustrate the operation of this plug-in, the picture \ref{fig:CodeRepetition1} shows two scripts containing same blocks with the only difference being received by parameter values.
+\paragraph{}On the one hand, it is very common that students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX}, where X is an incremental number.
+When the aumount of characters in a project increases is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging for possible errors.
+\paragraph{}On the other hand, it is also very common to see students repeat code in the same project, sometimes even in the same character programs.
+Thus, besides complicating the maintenance and updating of the code, the conceptss of abstraction and modularization, two key components of the development of computational thinking \cite{wing2008computational}, are not trained.
+\paragraph{}In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment \cite{boe2013hairball}, which in turn makes use of kurt\textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.
+\paragraph{}The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, which indicates that the user has not changed its name and has left the one the system assigns by default to every new character added to the project.
+\paragraph{}The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
+In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that differ only in the received values are considered equal.
+To illustrate the operation of this plug-in, the picture \ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the velues received by parameter.
 \begin{figure}
@@
 
-\paragraph{}--Sin retocar--These scripts to be analyzed by the plug-in and translated into \textit{tokens} would be considered equal, because your code would be translated to the next set of blocks:
+\paragraph{}When these scripts are analyzed by the plug-in and translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
 \begin{verbatim}
@@
 \end{verbatim}
-\paragraph{} --Sin retocar-- Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks receiving multiple values per parameter and reusing the block created 	in both programs, with this solution would involve improvements as regards the maintenance, updating and debugging code.
+\paragraph{}Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter and reusing the block created in both programs.
+This solution would involve improvements regarding maintenance, updating and debugging.
 
@@
 
-\paragraph{}--Sin retocar-- To test the functionality of the plug-ins developed and test how these behaviors that we detected in our work are also repeated in the projects repository Scratch, we proceeded to download 100 similar projects.
-As the possibility for users to define their own blocks, using block \textit{def\_block}, only available since the release of Scratch 2.0 downloaded projects are all after this date.
+\paragraph{}In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also repeated in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
+As the possibility for users to define their own blocks, using \textit{def\_block} block, is only available since the release of Scratch 2.0, downloaded projects were all created after that date.
 

### 2014-04-17 paperFIE_en.tex
Finding section refined

@@
 \section{Findings}
-\paragraph{}--Sin retocar--
-The data obtained after analyzing 100 projects downloaded observations confirm that our students are also very common in community projects.
-\paragraph{}--Sin retocar--Regarding the lack of habit of renaming the characters and do not use the default name assigned by the environment, the 79 \% of the projects analyzed contained at least one character without renaming the type \textit{SpriteX}.
-\paragraph{}--Sin retocar--As an example to illustrate the drawbacks associated with this malpractice shown in picture \ref{fig:SpriteNaming}  a script that is difficult to read as it is necessary to explore the area to find characters which character refers to a particular block, making it difficult to debug the code before any problems or improvements and even slows own programming for the student, you should memorize or check the number of object you want to perform an action.
+\paragraph{}Data obtained after analyzing 100 downloaded projects confirm that observations made with our students are also very common in the community projects.
+\paragraph{}Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, the 79 \% of the analyzed projects contained at least one \textit{SpriteX} character not renamed.
+\paragraph{}As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as it is necessary to explore the character area to find which character refers to a particular block, making it difficult to debug the code any problems should arise, making it hard to make any improvement in the code and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.
 \begin{figure}
@@
 \end{figure}
-\paragraph{}--Sin retocar--As for the repetition code, 62\% of the projects analyzed contained at least one repeated program.
-In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define a block itself and can be reused in other parts of the project.
+\paragraph{}As for the repetition code, 62\% of the analyzed projects contained at least one repeated program.
+In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project.
 

### 2014-04-17 paperFIE_en.tex
Conclussions section refined

@@
 
-\paragraph{} --Sin retocar--Regarding the named object is curious that other users Scratch named correctly, ie, significantly, the variables used in their projects.
-In our opinion this is because when you create a variable is mandatory to assign a name, because the environment does not name by default.
-However, when creating a new character is asinga named \textit{SpriteX} automatically.
-Maybe a change of environment that requires the user to name the new objects eliminate this malpractice.
-\paragraph{} --Sin retocar-- Regarding the code repetition and low usage users Scratch make the definition of own blocks (the equivalent of creating methods or procedures in other languages), you may work as performed by Seiter and Foreman \cite {seiter2013modeling} in which shows the capacity for abstraction and modularization are concepts that seem habilidadades develop from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 \footnote{http://scratch.mit.edu/statistics/\#age}.
-Future studies should further investigate this line to continue advancing in the definition of an appropriate framework for different ages and to check whether users online learning and those who do in a regulated environment show rates than use this functionality .
+\paragraph{}Regarding the named object it is curious that Scratch users name correctly, ie significantly, the variables used in their projects.
+In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default.
+However, when creating a new character a  \textit{SpriteX} name is assigned automatically.
+A change on this feature of the environment, obliguing users to name new objects, would eliminate this malpractice.
+\paragraph{}Regarding the code repetition and low usage Scratch users make of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old \footnote{http://scratch.mit.edu/statistics/\#age}.
+Future studies should further investigate this line to continue advancing in the definition of an appropriate framework for different ages and to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality .
 

### 2014-04-17 paperFIE_en.tex
Fix errata

@@
 
-\paragraph{}In our work as teachers with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits in a regular basis that have to do with named object and code repetition.
+\paragraph{}In our work as teachers with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits in a regular basis that have to do with object namig and code repetition.
 
@@
 
-\paragraph{}Regarding the named object it is curious that Scratch users name correctly, ie significantly, the variables used in their projects.
+\paragraph{}Regarding the object namig it is curious that Scratch users name correctly, ie significantly, the variables used in their projects.
 In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default.
@@
 \paragraph{}Regarding the code repetition and low usage Scratch users make of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old \footnote{http://scratch.mit.edu/statistics/\#age}.
-Future studies should further investigate this line to continue advancing in the definition of an appropriate framework for different ages and to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality .
+Future studies should further investigate this line to continue advancing in the definition of an appropriate framework for different ages and to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.
 

### 2014-04-17 paperFIE_en.tex
Fix erratas and misspelled words. Refine some expressions.

@@
 Scratch \cite{resnick2009scratch}  is a programming environment that includes a visual programming language designed for children from 6 years old, a development environment and a website where the community can host their projects; run, study and reuse other programs; and share ideas or suggestions with other programmers.
-Although there are several similar visual programming languages, Scratch is undoubtedly the most successful and the one that is being used more regularly worldwide, with more than three million registered users on its website and more than five million shared project in its repository \footnote{http://scratch.mit.edu/statistics/}.
+Although there are several similar visual programming languages, Scratch is undoubtedly the most successful and the one that is being used more regularly worldwide, with more than three million registered users on its website and more than five million shared projects in its repository \footnote{http://scratch.mit.edu/statistics/}.
 \paragraph{}Scratch is being used both in extracurricular activities \cite{maloney2008programming, kafai2010entering} or summer camps \cite{adams2010scratching, franklin2013assessment}, as in schools \cite{wilson2012evaluation}, high schools \cite{meerbaum2013learning} and even universities worldwide \cite{wolz2009starting, malan2007scratch}.
-Its capacity to bring computer programming to children and youth with the aim that in the future they would be willing to continue their studies in this discipline, and also its success to teach both basic and advanced programming concepts has been documented.
+Its capacity to bring computer programming to children and youth with the aim that in the future they would be willing to continue their studies in this discipline, and also its success to teach both basic and advanced programming concepts have been documented.
 \paragraph{}However, it is also possible to find studies that have detected some bad programming habits in students learning to program in this environment \cite{meerbaum2011habits}.
@@
 Scratch was designed with the main goal of encouraging young people to program, and most of its users are novice learners with no prior programming experience \cite{maloney2010scratch}.
-The environment is designed to encourage users to manipulate and modify various media types, so that children and youth can create projects that they find interesting and attractive as animated stories, games or interactive presentations \cite{maloney2008programming}
+The environment is designed to encourage users to manipulate and modify various media types, so that children and youth can create projects that they find interesting and attractive, such as animated stories, games or interactive presentations \cite{maloney2008programming}
 \paragraph{}Scratch also offers a website where the projects programmed by the community are published, the code of each program can be studied, and it is possible to make suggestions and reuse the code of any project, so there are lots of valuable resources available for anyone willing to start to program.
 This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community \cite{scaffidi2012skill}.
-\paragraph{}As mentioned in the introduction, many papers showing that Scratch works as a tool for learning to program can be found \cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, as there are success stories in different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
+\paragraph{}As mentioned in the introduction, many papers showing that Scratch works as a tool for learning to program can be found \cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, as there are success stories at different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
 \paragraph{}Despite these good results, Meerbaum-Salant, Armoni and Ben-Ari \cite{meerbaum2011habits} prove that students learning to program in this environment show some bad habits that are contrary to recommended programming practices.
@@
 \paragraph{} Aiming to simplify the process of evaluating the Computational Thinking and help evaluators with an automation tool, Hairball was created \cite{boe2013hairball}.
-Hairball is a static analyzer of Scratch projects inspired by Lint \footnote{http:// www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program or including code that is never run.
+Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{http:// www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program or including code that is never run.
 
@@
 \paragraph{}On the one hand, it is very common that students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX}, where X is an incremental number.
-When the aumount of characters in a project increases is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging for possible errors.
-\paragraph{}On the other hand, it is also very common to see students repeat code in the same project, sometimes even in the same character programs.
-Thus, besides complicating the maintenance and updating of the code, the conceptss of abstraction and modularization, two key components of the development of computational thinking \cite{wing2008computational}, are not trained.
-\paragraph{}In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment \cite{boe2013hairball}, which in turn makes use of kurt\textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.
+When the aumount of characters in a project increases, it is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging to fix bugs.
+\paragraph{}On the other hand, it is also very common to see students repeating code in the same project, sometimes even in the same character programs.
+Thus, besides complicating the maintenance and updating of the code, abstraction and modularization, two key components of the development of computational thinking \cite{wing2008computational}, are not trained.
+\paragraph{}In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment \cite{boe2013hairball}, which in turn makes use of \textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.
 \paragraph{}The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, which indicates that the user has not changed its name and has left the one the system assigns by default to every new character added to the project.
 \paragraph{}The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
-In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that differ only in the received values are considered equal.
-To illustrate the operation of this plug-in, the picture \ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the velues received by parameter.
+In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values the receive are considered to be equal.
+To illustrate the operation of this plug-in, picture \ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.
 \begin{figure}
@@
 
-\paragraph{}When these scripts are analyzed by the plug-in and translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
+\paragraph{}When these scripts are analyzed by the plug-in and are translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
 \begin{verbatim}
@@
 
-\paragraph{}In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also repeated in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
-As the possibility for users to define their own blocks, using \textit{def\_block} block, is only available since the release of Scratch 2.0, downloaded projects were all created after that date.
+\paragraph{}In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also found in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
+As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created after that date.
 
@@
 \paragraph{}Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, the 79 \% of the analyzed projects contained at least one \textit{SpriteX} character not renamed.
-\paragraph{}As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as it is necessary to explore the character area to find which character refers to a particular block, making it difficult to debug the code any problems should arise, making it hard to make any improvement in the code and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.
+\paragraph{}As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as the character area must be explored to find which character refers to a particular block, making the program difficult to debug any problems should arise, making it hard to code improvements and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.
 \begin{figure}
@@
 
-\paragraph{}Regarding the object namig it is curious that Scratch users name correctly, ie significantly, the variables used in their projects.
+\paragraph{}Regarding the object namig, it is curious that Scratch users name correctly, ie significantly, the variables used in their projects.
 In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default.
@@
 A change on this feature of the environment, obliguing users to name new objects, would eliminate this malpractice.
-\paragraph{}Regarding the code repetition and low usage Scratch users make of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old \footnote{http://scratch.mit.edu/statistics/\#age}.
-Future studies should further investigate this line to continue advancing in the definition of an appropriate framework for different ages and to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.
+\paragraph{}Regarding the code repetition and low usage Scratch users make of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}.
+Future studies should further investigate to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.
 

### 2014-04-18 paperFIE_en.tex
Refine some expressions

@@
 However, when creating a new character a  \textit{SpriteX} name is assigned automatically.
-A change on this feature of the environment, obliguing users to name new objects, would eliminate this malpractice.
-\paragraph{}Regarding the code repetition and low usage Scratch users make of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}.
+In our views, a change on this feature of the environment, obliguing users to name new objects, might eliminate this malpractice.
+\paragraph{}Regarding the code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}.
 Future studies should further investigate to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.

### 2014-04-18 paperFIE_en.tex
Fix errata and change a couple of words

@@
 \paragraph{}The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
-In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values the receive are considered to be equal.
+In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal.
 To illustrate the operation of this plug-in, picture \ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.
@@
 
-\paragraph{}In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also found in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
+\paragraph{}In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
 As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created after that date.

### 2014-04-18 paperFIE_en.tex
Some minor changes to the paper text. Adding format files

@@
-\documentclass[a4paper,10pt]{article}
+%\documentclass[a4paper,10pt]{article}
+\documentclass[conference]{format/IEEEtran}
 %\usepackage[utf8x]{inputenc}
@@
 %opening
-\title{Automatic detection of bad habits of programming in Scratch}
-\author{Jesús Moreno, Gregorio Robles}
+\title{Automatic Detection of Bad Programming Habits in Scratch: A Preliminary Study}
+%\author{Jesús Moreno, Gregorio Robles}
+
+\author{\IEEEauthorblockN{Jesús Moreno}
+\IEEEauthorblockA{INTEF \& Programamos.es\\
+Madrid, Spain\\
+j.morenol@gmail.com}
+\and
+\IEEEauthorblockN{Gregorio Robles}
+\IEEEauthorblockA{GSyC/LibreSoft\\Universidad Rey Juan Carlos\\
+Madrid, Spain\\
+grex@gsyc.urjc.es}}
 
@@
 \begin{abstract}
-This paper describes a method for automatic detection of two very bad programming habits present in students learning to program with Scratch, such as the repetition of code and object naming, and raises ideas to try to avoid such situations.
+This paper describes a method for automatic detection of two bad programming habits that students learning to program with Scratch incur, such as the repetition of code and object naming, and raises ideas to try to avoid such situations.
+FIXME: some methodology, results and conclusions should be included
 
@@
 \section{Introduction}
-Scratch \cite{resnick2009scratch}  is a programming environment that includes a visual programming language designed for children from 6 years old, a development environment and a website where the community can host their projects; run, study and reuse other programs; and share ideas or suggestions with other programmers.
-Although there are several similar visual programming languages, Scratch is undoubtedly the most successful and the one that is being used more regularly worldwide, with more than three million registered users on its website and more than five million shared projects in its repository \footnote{http://scratch.mit.edu/statistics/}.
-\paragraph{}Scratch is being used both in extracurricular activities \cite{maloney2008programming, kafai2010entering} or summer camps \cite{adams2010scratching, franklin2013assessment}, as in schools \cite{wilson2012evaluation}, high schools \cite{meerbaum2013learning} and even universities worldwide \cite{wolz2009starting, malan2007scratch}.
-Its capacity to bring computer programming to children and youth with the aim that in the future they would be willing to continue their studies in this discipline, and also its success to teach both basic and advanced programming concepts have been documented.
-\paragraph{}However, it is also possible to find studies that have detected some bad programming habits in students learning to program in this environment \cite{meerbaum2011habits}.
-In this line, in our work as teachers we have detected other bad habits that, despite insisting our students to try to avoid them, we see how they are constantly repeated by pupils learning to program in Scratch.
-As these habits are contrary to the basic programming recommendations, our study tries to detect if they are also common in the projects shared on the Scratch repository and propose some ideas to try to avoid such situations.
-\paragraph{}In section 2 a brief overview of Scratch is presented, works that study the learning demonstrated by students learning to program in this environment are reviewed, and studies that automate the evaluation of this learning are exposed.
-In section 3, our work methodology is described.
-Section 4 presents the obtained results.
-Finally, in section 5 the conclusions of our study are discussed and some ideas and suggestions for future work are provided.
+
+Scratch~\cite{resnick2009scratch}  is a programming environment including a visual programming language designed for children over 6 years old, a development environment and a website where the community can host their projects, run, study and reuse other programs, and share ideas or suggestions with other programmers.
+Although there are several similar visual programming languages, Scratch is undoubtedly the most successful one with more than three million registered users on its website and more than five million shared projects in its repository\footnote{See http://scratch.mit.edu/statistics/}.
+
+Scratch is being used both in extracurricular activities~\cite{maloney2008programming, kafai2010entering} or summer camps~\cite{adams2010scratching, franklin2013assessment}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{wolz2009starting, malan2007scratch}.
+Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts have been documented (FIXME: reference required).
+
+However, it is also possible to find studies that have detected some bad programming habits in students learning to program in this environment~\cite{meerbaum2011habits}.
+In this line, in our work as instructors we have detected bad habits despite having insisted our students to avoid them.
+As these habits are contrary to the basic programming recommendations, our study tries to detect if they are also common in the projects shared in the Scratch repository, and propose some ideas to try to avoid such situations.
+
+The structure of this paper is as follows: In section~\ref{sec:background} a brief overview of Scratch is presented, including some research studies that have investigated the learning in this environment, and studies that automate the evaluation of this learning are exposed.
+In section~\ref{sec:methodology}, our work methodology is described.
+Section~\ref{sec:findings} presents the obtained results.
+Finally, in Section~\ref{sec:conclusions} the conclusions of our study are discussed and some ideas and suggestions for future work are provided.
 
 \section{Background}
+\label{sec:background}
 
-\paragraph{}Scratch is a visual programming environment developed by the Lifelong Kindergarten group at MIT Media Laboratory that allows to program interactive projects using a drag-and-drop approach with different pieces or \textit {Lego} style blocks \cite{resnick2009scratch}.
-Scratch was designed with the main goal of encouraging young people to program, and most of its users are novice learners with no prior programming experience \cite{maloney2010scratch}.
-The environment is designed to encourage users to manipulate and modify various media types, so that children and youth can create projects that they find interesting and attractive, such as animated stories, games or interactive presentations \cite{maloney2008programming}
-\paragraph{}Scratch also offers a website where the projects programmed by the community are published, the code of each program can be studied, and it is possible to make suggestions and reuse the code of any project, so there are lots of valuable resources available for anyone willing to start to program.
-This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community \cite{scaffidi2012skill}.
-\paragraph{}As mentioned in the introduction, many papers showing that Scratch works as a tool for learning to program can be found \cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, as there are success stories at different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
-\paragraph{}Despite these good results, Meerbaum-Salant, Armoni and Ben-Ari \cite{meerbaum2011habits} prove that students learning to program in this environment show some bad habits that are contrary to recommended programming practices.
-In particular, they show excessive decomposition tendency and a development process absolutly from bottom to top starting with individual blocks of Scratch.
-\paragraph{} Aiming to simplify the process of evaluating the Computational Thinking and help evaluators with an automation tool, Hairball was created \cite{boe2013hairball}.
+Scratch is a visual programming environment developed by the Lifelong Kindergarten group at MIT Media Laboratory that allows to program interactive projects using a drag-and-drop approach with different pieces or \textit {Lego} style blocks~\cite{resnick2009scratch}.
+Scratch was designed with the main goal of encouraging young people to program, and most of its users are novice learners with no prior programming experience~\cite{maloney2010scratch}.
+The environment is designed to encourage users to manipulate and modify various media types, so that children and youth can create projects that they find interesting and attractive, such as animated stories, games or interactive presentations~\cite{maloney2008programming}.
+
+Scratch also offers a website where the projects programmed by the community are published, the code of each program can be studied, and it is possible to make suggestions and reuse the code of any project, so there are lots of valuable resources available for anyone willing to start to program.
+This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community~\cite{scaffidi2012skill}.
+
+As mentioned in the introduction, many papers showing that Scratch works as a tool for learning to program can be found~\cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, as there are success stories at different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
+
+Despite these good results, Meerbaum-Salant, Armoni and Ben-Ari~\cite{meerbaum2011habits} prove that students learning to program in this environment show some bad habits that are contrary to recommended programming practices.
+In particular, they show excessive decomposition tendency and a development process absolutely from bottom to top starting with individual blocks of Scratch.
+
+ Aiming to simplify the process of evaluating the Computational Thinking and help evaluators with an automation tool, Hairball was created~\cite{boe2013hairball}.
 Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{http:// www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program or including code that is never run.
@@
 \section{Methodology}
+\label{sec:methodology}
 
-\paragraph{}In our work as teachers with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits in a regular basis that have to do with object namig and code repetition.
+In our work as instructors with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits in a regular basis that have to do with object naming and code repetition.
 
-\paragraph{}On the one hand, it is very common that students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX}, where X is an incremental number.
-When the aumount of characters in a project increases, it is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging to fix bugs.
-\paragraph{}On the other hand, it is also very common to see students repeating code in the same project, sometimes even in the same character programs.
-Thus, besides complicating the maintenance and updating of the code, abstraction and modularization, two key components of the development of computational thinking \cite{wing2008computational}, are not trained.
-\paragraph{}In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment \cite{boe2013hairball}, which in turn makes use of \textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.
-\paragraph{}The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, which indicates that the user has not changed its name and has left the one the system assigns by default to every new character added to the project.
-\paragraph{}The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
+On the one hand, it is very common that students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX}, where X is an incremental number.
+When the amount of characters in a project increases, it is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging to fix bugs.
+
+On the other hand, it is also very common to see students repeating code in the same project, sometimes even in the same character programs.
+Thus, besides complicating the maintenance and updating of the code, abstraction and modularization, two key components of the development of computational thinking~\cite{wing2008computational}, are not trained.
+
+In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment~\cite{boe2013hairball}, which in turn makes use of \textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.
+
+The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, which indicates that the user has not changed its name and has left the one the system assigns by default to every new character added to the project.
+
+The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
 In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal.
@@
 
-\paragraph{}When these scripts are analyzed by the plug-in and are translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
+When these scripts are analyzed by the plug-in and are translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
 \begin{verbatim}
@@
 \end{verbatim}
-\paragraph{}Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter and reusing the block created in both programs.
+
+Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter and reusing the block created in both programs.
 This solution would involve improvements regarding maintenance, updating and debugging.
@@
 
-\paragraph{}In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
+In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
 As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created after that date.
@@
 \section{Findings}
-\paragraph{}Data obtained after analyzing 100 downloaded projects confirm that observations made with our students are also very common in the community projects.
-\paragraph{}Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, the 79 \% of the analyzed projects contained at least one \textit{SpriteX} character not renamed.
-\paragraph{}As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as the character area must be explored to find which character refers to a particular block, making the program difficult to debug any problems should arise, making it hard to code improvements and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.
+\label{sec:findings}
+
+Data obtained after analyzing 100 downloaded projects confirm that observations made with our students are also very common in the community projects.
+
+Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, the 79\% of the analyzed projects contained at least one \textit{SpriteX} character not renamed.
+
+As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as the character area must be explored to find which character refers to a particular block, making the program difficult to debug any problems should arise, making it hard to code improvements and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.
+
 \begin{figure}
@@
 \end{figure}
-\paragraph{}As for the repetition code, 62\% of the analyzed projects contained at least one repeated program.
+
+As for the repetition code, 62\% of the analyzed projects contained at least one repeated program.
 In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project.
 
-\section{Conclussions}
+\section{Conclusions and further research}
+\label{sec:conclusions}
 
-\paragraph{}Regarding the object namig, it is curious that Scratch users name correctly, ie significantly, the variables used in their projects.
+FIXME: In this paper... short summary
+
+Regarding the object naming, it is curious that Scratch users name correctly, i.e., semantically meaningful, the variables used in their projects.
 In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default.
 However, when creating a new character a  \textit{SpriteX} name is assigned automatically.
-In our views, a change on this feature of the environment, obliguing users to name new objects, might eliminate this malpractice.
-\paragraph{}Regarding the code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman \cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}.
+In our views, a change of this feature of the environment, forcing users to name new objects, might eliminate this bad practice.
+
+Regarding the code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman~\cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}.
 Future studies should further investigate to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.
 
-\newpage
+We plan to... FIXME
+
+\section*{Acknowledgments}
+
+The work of Gregorio Robles has
+been funded in part by the Spanish Government under project 
+SobreSale (TIN2011-28110) and by the Comunidad de Madrid 
+under project eMadrid (FIXME).
+
+\bibliographystyle{abbrv}
 \bibliography{scratch}
-\bibliographystyle{unsrt}
 \end{document}

### 2014-04-18 paperFIE_en.tex
Fixing the figures and adding some more FIXMEs to be addressed

@@
 This paper describes a method for automatic detection of two bad programming habits that students learning to program with Scratch incur, such as the repetition of code and object naming, and raises ideas to try to avoid such situations.
-FIXME: some methodology, results and conclusions should be included
+FIXME: some methodology, results and conclusions should be included in the abstract.
 
@@
 Scratch is being used both in extracurricular activities~\cite{maloney2008programming, kafai2010entering} or summer camps~\cite{adams2010scratching, franklin2013assessment}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{wolz2009starting, malan2007scratch}.
-Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts have been documented (FIXME: reference required).
+Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been documented (FIXME: reference required).
 
@@
 
-The structure of this paper is as follows: In section~\ref{sec:background} a brief overview of Scratch is presented, including some research studies that have investigated the learning in this environment, and studies that automate the evaluation of this learning are exposed.
-In section~\ref{sec:methodology}, our work methodology is described.
-Section~\ref{sec:findings} presents the obtained results.
-Finally, in Section~\ref{sec:conclusions} the conclusions of our study are discussed and some ideas and suggestions for future work are provided.
+FIXME: the goal of this paper is to...
+
+The structure of this paper is as follows: In section~\ref{sec:background} a brief overview of Scratch is presented, including some research studies that have investigated learning in this environment and other studies that automate the evaluation of learning.
+In section~\ref{sec:methodology} our work methodology is briefly described.
+Section~\ref{sec:findings} presents the results we have obtained from applying our methodology on a small sample of Scratch projects.
+Finally, Section~\ref{sec:conclusions} contains the conclusions of our study and some ideas and suggestions for future work are discussed.
 
@@
 
-Scratch is a visual programming environment developed by the Lifelong Kindergarten group at MIT Media Laboratory that allows to program interactive projects using a drag-and-drop approach with different pieces or \textit {Lego} style blocks~\cite{resnick2009scratch}.
+Scratch is a visual programming environment, developed by the Lifelong Kindergarten group at the MIT Media Laboratory, that allows to program interactive projects using a drag-and-drop approach with different pieces or \textit {Lego} style blocks~\cite{resnick2009scratch}.
 Scratch was designed with the main goal of encouraging young people to program, and most of its users are novice learners with no prior programming experience~\cite{maloney2010scratch}.
-The environment is designed to encourage users to manipulate and modify various media types, so that children and youth can create projects that they find interesting and attractive, such as animated stories, games or interactive presentations~\cite{maloney2008programming}.
+The environment is designed to encourage users to manipulate and modify various media types, so that children and teenagers can create projects that they find interesting and attractive, such as animated stories, games or interactive presentations~\cite{maloney2008programming}.
 
-Scratch also offers a website where the projects programmed by the community are published, the code of each program can be studied, and it is possible to make suggestions and reuse the code of any project, so there are lots of valuable resources available for anyone willing to start to program.
+Scratch also offers a website where the projects programmed by the community are published.
+The code of programs hosted there can be studied and analyzed, and it is possible to make suggestions and reuse the code, so there are lots of valuable resources available for anyone willing to start to program.
 This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community~\cite{scaffidi2012skill}.
 
-As mentioned in the introduction, many papers showing that Scratch works as a tool for learning to program can be found~\cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, as there are success stories at different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
+% repe
+%There are many papers showing that Scratch works as a tool for learning to program~\cite{maloney2008programming, franklin2013assessment, scaffidi2012skill, malan2007scratch}, and there are success stories at different educational levels, environments and disciplines showing that students learn concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation.
 
-Despite these good results, Meerbaum-Salant, Armoni and Ben-Ari~\cite{meerbaum2011habits} prove that students learning to program in this environment show some bad habits that are contrary to recommended programming practices.
-In particular, they show excessive decomposition tendency and a development process absolutely from bottom to top starting with individual blocks of Scratch.
+Despite the positive impact of Scratch, Meerbaum-Salant, Armoni and Ben-Ari~\cite{meerbaum2011habits} prove that students learning to program in this environment show some bad habits that are contrary to recommended programming practices.
+In particular, they show a tendency to excessive decomposition and a bottom-up development process, starting with individual blocks of Scratch.
 
- Aiming to simplify the process of evaluating the Computational Thinking and help evaluators with an automation tool, Hairball was created~\cite{boe2013hairball}.
-Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{http:// www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program or including code that is never run.
+Aiming to simplify the process of evaluating computational thinking and help evaluators with an automatic tool, Hairball was created~\cite{boe2013hairball}.
+Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{http://www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program, including code that is never run, among others.
 
@@
   \centering
-    \includegraphics{img/CodeRepetition1.png}
+    \includegraphics[width=9cm]{img/CodeRepetition1.png}
   \caption{Two scripts repeating code}
@@
   \centering
-    \includegraphics{img/CodeRepetition2.png}
+    \includegraphics[width=9cm]{img/CodeRepetition2.png}
   \caption{Defining own blocks to avoid code repetition}
@@
   \centering
-    \includegraphics{img/SpriteNaming.png}
+    \includegraphics[width=9cm]{img/SpriteNaming.png}
   \caption{Sprites with default names}
@@
 
-FIXME: In this paper... short summary
+FIXME: In this paper... short summary of what we have done and why
 
@@
 
-We plan to... FIXME
+In the near future, we plan to... FIXME... talk about next steps: i) more analysis, ii) more projects, iii) web-based tool that provides learning-oriented feedback in an automatic way.
 

### 2014-04-19 paperFIE_en.tex
Conclussions: add a short summary of the work done and plans for near future

@@
 Madrid, Spain\\
-j.morenol@gmail.com}
+jesus.moreno@programamos.es}
 \and
@@
 
-Scratch is being used both in extracurricular activities~\cite{maloney2008programming, kafai2010entering} or summer camps~\cite{adams2010scratching, franklin2013assessment}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{wolz2009starting, malan2007scratch}.
-Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been documented (FIXME: reference required).
+Scratch is being used both in extracurricular activities~\cite{kafai2010entering} or summer camps~\cite{adams2010scratching}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{malan2007scratch}.
+Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been documented~\cite{maloney2008programming, franklin2013assessment, wolz2009starting}.
 
@@
 
-FIXME: In this paper... short summary of what we have done and why
+%FIXME: In this paper... short summary of what we have done and why
+
+This paper shows the preliminary results of a study regarding two bad programming habits we have detected in our work as instructors with high school students learning to program with Scratch.
+In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 proyects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected proyects, 79\% and 62\%  respectively, fell into these issues.
 
@@
 
-In the near future, we plan to... FIXME... talk about next steps: i) more analysis, ii) more projects, iii) web-based tool that provides learning-oriented feedback in an automatic way.
+%In the near future, we plan to... FIXME... talk about next steps: i) more analysis, ii) more projects, iii) web-based tool that provides learning-oriented feedback in an automatic way.
+
+In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and analyzing a much larger number of projects.
+In this sense, the release of a research dataset with the first five years of data from the Scratch repository\footnote{https://sites.google.com/site/scratchdatameeting/} will be certainly a great help to our work.
+Finally, we are in the process of developing a website that will allow students to receive feedback from each of its programs with warnings of potential errors and tips to improve their code, and will offer teachers a tool to assist them in correcting their students proyects.
 

### 2014-04-19 paperFIE_en.tex
Extend abstract and modify some expressions.

@@
 \begin{abstract}
-This paper describes a method for automatic detection of two bad programming habits that students learning to program with Scratch incur, such as the repetition of code and object naming, and raises ideas to try to avoid such situations.
+%This paper describes a method for automatic detection of two bad programming habits that students learning to program with Scratch incur, such as the repetition of code and object naming, and raises ideas to try to avoid such situations.
 FIXME: some methodology, results and conclusions should be included in the abstract.
+Using Scratch as a tool to teach computing science or to work computational thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented success case studies from primary school to university.
+
+However, there are reports indicating that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations.
+
+In our work as instructors with high school students learning to program with Scratch, we also found some common bad practices, such as the repetition of code and object naming, and this study focuses on checking whether these habits are repeated in the projects shared on the community website.
+
+To test this hypothesys, we downloaded 100 projects from the Scratch website and analyzed them with two plug-ins we developed to detect these bad practices extending the functionality of the tool Hairball, a static code analyzer for Scratch projects.
+The obtained results indicate that, in general, the projects in the repository also incur in these type of malpractices.
+
+Along with suggestions for future work, some ideas that might assist to avoid such situations are proposed in the conclusions of the paper.
 
@@
 In this line, in our work as instructors we have detected bad habits despite having insisted our students to avoid them.
-As these habits are contrary to the basic programming recommendations, our study tries to detect if they are also common in the projects shared in the Scratch repository, and propose some ideas to try to avoid such situations.
+As these habits are contrary to the basic programming recommendations, our study tries to find out if they are also common in the projects shared in the Scratch repository, and propose some ideas to try to avoid such situations.
 
 FIXME: the goal of this paper is to... 
+% ¿No estaría explicado justo en el párrafo de encima? ¿Habría que explicarlo más detalladamente?
 
@@
 
-In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and analyzing a much larger number of projects.
+In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and analyzing a much larger number of projects to achieve more accurate conclusions.
 In this sense, the release of a research dataset with the first five years of data from the Scratch repository\footnote{https://sites.google.com/site/scratchdatameeting/} will be certainly a great help to our work.

### 2014-04-19 paperFIE_en.tex
Modify abstract

@@
 FIXME: some methodology, results and conclusions should be included in the abstract.
-Using Scratch as a tool to teach computing science or to work computational thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented success case studies from primary school to university.
+Using Scratch as a tool to teach Computer Science or develop Computational Thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented success case studies from primary school to university experiences.
 
 However, there are reports indicating that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations.
+In our work as instructors with high school students learning to program with Scratch, we also detected some common bad practices, such as the repetition of code and object naming, and this study focuses on checking whether these issues can be found in the projects shared on the community website.
 
-In our work as instructors with high school students learning to program with Scratch, we also found some common bad practices, such as the repetition of code and object naming, and this study focuses on checking whether these habits are repeated in the projects shared on the community website.
-
-To test this hypothesys, we downloaded 100 projects from the Scratch website and analyzed them with two plug-ins we developed to detect these bad practices extending the functionality of the tool Hairball, a static code analyzer for Scratch projects.
-The obtained results indicate that, in general, the projects in the repository also incur in these type of malpractices.
+To test this hypothesys, we downloaded 100 projects from the Scratch website and analyzed them with two plug-ins we developed to automatically detect these bad practices.
+The plug-ins extend the functionality of the tool Hairball, a static code analyzer for Scratch projects.
+The obtained results indicate that, in general, projects in the repository also incur in the investigated malpractices.
 
@@
 
-In order to test the functionality of the developed plug-ins and test whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
+In order to test the functionality of the developed plug-ins and check whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
 As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created after that date.

### 2014-04-20 paperFIE_en.tex
Introduction: add pararaph with the goal of this paper. Modify some expressions.

@@
 FIXME: some methodology, results and conclusions should be included in the abstract.
-Using Scratch as a tool to teach Computer Science or develop Computational Thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented success case studies from primary school to university experiences.
+Using Scratch as a tool to teach Computer Science or develop Computational Thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented case studies from primary school to university.
 
 However, there are reports indicating that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations.
-In our work as instructors with high school students learning to program with Scratch, we also detected some common bad practices, such as the repetition of code and object naming, and this study focuses on checking whether these issues can be found in the projects shared on the community website.
+In our work as instructors with high school students learning to program with Scratch, we have detected other bad practices, such as the repetition of code and object naming, that are repeated on a regular basis, and this study focuses on checking whether these issues are also present at the projects shared on the community website.
 
 To test this hypothesys, we downloaded 100 projects from the Scratch website and analyzed them with two plug-ins we developed to automatically detect these bad practices.
-The plug-ins extend the functionality of the tool Hairball, a static code analyzer for Scratch projects.
+The plug-ins extend the functionality of the tool Hairball, a static code analyzer of Scratch projects.
+
 The obtained results indicate that, in general, projects in the repository also incur in the investigated malpractices.
-
 Along with suggestions for future work, some ideas that might assist to avoid such situations are proposed in the conclusions of the paper.
@@
 However, it is also possible to find studies that have detected some bad programming habits in students learning to program in this environment~\cite{meerbaum2011habits}.
-In this line, in our work as instructors we have detected bad habits despite having insisted our students to avoid them.
-As these habits are contrary to the basic programming recommendations, our study tries to find out if they are also common in the projects shared in the Scratch repository, and propose some ideas to try to avoid such situations.
+In this line, in our work as instructors we have also found some bad habits despite having insisted our students to avoid them.
+As these habits are contrary to the basic programming recommendations, our study tries to find out if they are also common in the projects shared in the Scratch repository.
 
-FIXME: the goal of this paper is to... 
-% ¿No estaría explicado justo en el párrafo de encima? ¿Habría que explicarlo más detalladamente?
+%FIXME: the goal of this paper is to... 
+The goal of this paper, therefore, is to determine whether these bad habits are common practices in the Scratch community and propose some ideas to try to avoid such situations.
+In order to assist evaluators to detect these malpractices we developed two plug-ins that automate their detection by extending the functionality of Hairball, a static code analyzer of Scratch projects.
 
@@
 
-In our work as instructors with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits in a regular basis that have to do with object naming and code repetition.
+In our work as instructors with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits on a regular basis that have to do with object naming and code repetition.
 
@@
 This paper shows the preliminary results of a study regarding two bad programming habits we have detected in our work as instructors with high school students learning to program with Scratch.
-In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 proyects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected proyects, 79\% and 62\%  respectively, fell into these issues.
+In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 proyects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected proyects, 79\% and 62\%  respectively, fall into these issues.
 

### 2014-04-23 paperFIE_en.tex
Reviewing the paper. Changes here and there

@@
 \usepackage{graphicx}
+\usepackage{url}
 
@@
 \author{\IEEEauthorblockN{Jesús Moreno}
-\IEEEauthorblockA{INTEF \& Programamos.es\\
+\IEEEauthorblockA{Programamos.es \& Instituto Nacional de \\ Tecnologías Educativas y de Formación del Profesorado\\
 Madrid, Spain\\
@@
 FIXME: some methodology, results and conclusions should be included in the abstract.
-Using Scratch as a tool to teach Computer Science or develop Computational Thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented case studies from primary school to university.
+Using the Scratch environment as a tool to teach programming skills or develop computational thinking is increasingly common in all levels of education, well-documented case studies from primary school to university can be found.
 
-However, there are reports indicating that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations.
-In our work as instructors with high school students learning to program with Scratch, we have detected other bad practices, such as the repetition of code and object naming, that are repeated on a regular basis, and this study focuses on checking whether these issues are also present at the projects shared on the community website.
+However, there are reports that indicate that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations.
+In our work as instructors with high school students, we have detected these and other bad practices, such as the repetition of code and object naming, on a regular basis.
+This paper focuses on verifying whether these issues can be generally be found in the Scratch community, by analyzing a large number of projects available on the Scractch community website.
 
-To test this hypothesys, we downloaded 100 projects from the Scratch website and analyzed them with two plug-ins we developed to automatically detect these bad practices.
-The plug-ins extend the functionality of the tool Hairball, a static code analyzer of Scratch projects.
+To test this hypothesis, we downloaded 100 projects and analyzed them with two plug-ins we developed to automatically detect these bad practices.
+The plug-ins extend the functionality of the Hairball tool, a static code analyzer for Scratch projects.
 
-The obtained results indicate that, in general, projects in the repository also incur in the investigated malpractices.
-Along with suggestions for future work, some ideas that might assist to avoid such situations are proposed in the conclusions of the paper.
-
+The results obtained show that, in general, projects in the repository also incur in the investigated malpractices.
+Along with suggestions for future work, some ideas that might assist to address such situations are proposed in the conclusions of the paper.
 \end{abstract}
@@
 Scratch~\cite{resnick2009scratch}  is a programming environment including a visual programming language designed for children over 6 years old, a development environment and a website where the community can host their projects, run, study and reuse other programs, and share ideas or suggestions with other programmers.
-Although there are several similar visual programming languages, Scratch is undoubtedly the most successful one with more than three million registered users on its website and more than five million shared projects in its repository\footnote{See http://scratch.mit.edu/statistics/}.
+Although there are several similar visual programming languages, Scratch is undoubtedly the most successful one with more than three million registered users in its website and more than five million shared projects in its repository\footnote{See \url{http://scratch.mit.edu/statistics/}}.
 
 Scratch is being used both in extracurricular activities~\cite{kafai2010entering} or summer camps~\cite{adams2010scratching}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{malan2007scratch}.
-Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been documented~\cite{maloney2008programming, franklin2013assessment, wolz2009starting}.
+Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been well documented~\cite{maloney2008programming, franklin2013assessment, wolz2009starting}.
 
@@
 %FIXME: the goal of this paper is to... 
-The goal of this paper, therefore, is to determine whether these bad habits are common practices in the Scratch community and propose some ideas to try to avoid such situations.
-In order to assist evaluators to detect these malpractices we developed two plug-ins that automate their detection by extending the functionality of Hairball, a static code analyzer of Scratch projects.
+The goal of this paper, therefore, is to determine whether these bad habits are common practices in the Scratch community in an automatic manner and propose some ideas to try to avoid such situations.
+In order to assist evaluators to detect these malpractices we have developed two plug-ins that automate the detection of the bad programming practices by extending the functionality of Hairball, a static code analyzer of Scratch projects.
 
@@
 Scratch also offers a website where the projects programmed by the community are published.
-The code of programs hosted there can be studied and analyzed, and it is possible to make suggestions and reuse the code, so there are lots of valuable resources available for anyone willing to start to program.
-This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community~\cite{scaffidi2012skill}.
+The code of programs hosted there can be run, and following the community idea of the free software movement it can be studied and analyzed.
+So, it is possible to make suggestions and to reuse the code, and there are lots of valuable resources available for anyone willing to start to program.
+This approach offers as well the opportunity to learn social skills related to programming and beyond, such as sharing and contributing to the community~\cite{scaffidi2012skill}.
 
@@
 Aiming to simplify the process of evaluating computational thinking and help evaluators with an automatic tool, Hairball was created~\cite{boe2013hairball}.
-Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{http://www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program, including code that is never run, among others.
+Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{\url{http://www.unix.com/man-page/FreeBSD/1/lint}} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program, including code that is never run, among others.
 
@@
 
-In our work as instructors with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits on a regular basis that have to do with object naming and code repetition.
+In our work as instructors with high school students, we found that, despite our constant insistence and warnings, many students show on a regular basis two bad habits that have to do with object naming and code repetition:
 
-On the one hand, it is very common that students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX}, where X is an incremental number.
-When the amount of characters in a project increases, it is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging to fix bugs.
+\begin{enumerate}
 
-On the other hand, it is also very common to see students repeating code in the same project, sometimes even in the same character programs.
-Thus, besides complicating the maintenance and updating of the code, abstraction and modularization, two key components of the development of computational thinking~\cite{wing2008computational}, are not trained.
+  \item Students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX} where X is an incremental number.
+When the amount of characters in a project is large, it becomes very hard to know what object relates to a given statement, contributing to poor code readability, slowing the project programming, and complicating debugging to fix bugs.
 
-In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment~\cite{boe2013hairball}, which in turn makes use of \textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.
+  \item Students repeat code in the same project, sometimes even in the same character programs.
+Thus, abstraction and modularization, two key components of the development of computational thinking~\cite{wing2008computational}, are not trained.
+And the \emph{code base} becomes more complicated to maintain and update in the future.
 
-The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, which indicates that the user has not changed its name and has left the one the system assigns by default to every new character added to the project.
+\end{enumerate}
 
-The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project.
-In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal.
-To illustrate the operation of this plug-in, picture \ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.
+In order to detect these two behaviors in Scratch projects automatically, we have developed following plug-ins for the Hairball environment~\cite{boe2013hairball}, which in turn make use of \textit{kurt}\footnote{\url{https://github.com/blob8108/kurt}}, a library implemented in Python.
+The developed plug-ins are:
+
+\begin{enumerate}
+
+  \item \textit{convention.SpriteNaming}\footnote{\url{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py}} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, indicative for the user not having changed the default name.
+
+As an example to illustrate the drawbacks associated with this bad programming behavior, Fig.~\ref{fig:SpriteNaming} shows a script that is difficult to read because the complete character area has to be explored to find what character refers to a particular block.
+This makes the program difficult to debug should errors arise, making it hard to improve the code and slowing down the programming tasks.
+
+\begin{figure}
+  \centering
+    \includegraphics[width=9cm]{img/SpriteNaming.png}
+  \caption{Sprites with default names}
+  \label{fig:SpriteNaming}
+\end{figure}
+
+  \item \textit{duplicate.DuplicateScripts}\footnote{\url{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py}} analyzes a Scratch project to try to locate repeated complete programs within a project.
+In order to do so, the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal.
+To illustrate the operation of this plug-in, Fig.~\ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.
+
 \begin{figure}
@@
 
-When these scripts are analyzed by the plug-in and are translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
+The Scratch scripts are analyzed by the plug-in and translated into \textit{tokens}.
+The analysis would result in considering them equal, as their code would be translated in both cases into the following sequence of \textit{tokens}:
 \begin{verbatim}
@@
 
-Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter and reusing the block created in both programs.
-This solution would involve improvements regarding maintenance, updating and debugging.
+Hence, the correct way to implement this functionality would have been as shown in Fig.~\ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter, and then reusing the block created in both programs.
+This solution would facilitate future maintenance, updating and debugging.
 
@@
     \includegraphics[width=9cm]{img/CodeRepetition2.png}
-  \caption{Defining own blocks to avoid code repetition}
+  \caption{Definition of blocks to avoid code repetition.}
   \label{fig:CodeRepetition2}
@@
 
-In order to test the functionality of the developed plug-ins and check whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository.
-As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created after that date.
+\end{enumerate}
+
+In order to test the functionality of the developed plug-ins and to verify whether these behaviors are common in projects found in the Scratch repository, we proceeded to randomly download 100 projects from the repository.
+As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created for this version.
+A replication package of the projects under study, and the specific version of the plug-ins used in it can be obtained publicly\footnote{FIXME: URL \url{http://gsyc.urjc.es/~grex/repro/2014-fie-scratch}}.
 
@@
 
-Data obtained after analyzing 100 downloaded projects confirm that observations made with our students are also very common in the community projects.
+The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects.
 
-Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, the 79\% of the analyzed projects contained at least one \textit{SpriteX} character not renamed.
-
-As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as the character area must be explored to find which character refers to a particular block, making the program difficult to debug any problems should arise, making it hard to code improvements and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.
-
-\begin{figure}
-  \centering
-    \includegraphics[width=9cm]{img/SpriteNaming.png}
-  \caption{Sprites with default names}
-  \label{fig:SpriteNaming}
-\end{figure}
+Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, 79\% of the analyzed projects contained at least one non-renamed \textit{SpriteX} character.
 
@@
 In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project.
+
+FIXME: here comes the table with some more results. Important: to note how the distribution (talk about mean and median).
 
@@
 This paper shows the preliminary results of a study regarding two bad programming habits we have detected in our work as instructors with high school students learning to program with Scratch.
-In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 proyects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected proyects, 79\% and 62\%  respectively, fall into these issues.
+In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 projects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected projects, 79\% and 62\%  respectively, fall into these issues.
 
-Regarding the object naming, it is curious that Scratch users name correctly, i.e., semantically meaningful, the variables used in their projects.
+Regarding object naming, it is worth to note that Scratch users name the variables used in their projects correctly, i.e., semantically meaningful.
 In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default.
@@
 
-Regarding the code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman~\cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}.
-Future studies should further investigate to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.
+Regarding code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman~\cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years\footnote{\url{http://scratch.mit.edu/statistics/#age}}.
+Future studies should further investigate how to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.
 
@@
 
-In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and analyzing a much larger number of projects to achieve more accurate conclusions.
-In this sense, the release of a research dataset with the first five years of data from the Scratch repository\footnote{https://sites.google.com/site/scratchdatameeting/} will be certainly a great help to our work.
-Finally, we are in the process of developing a website that will allow students to receive feedback from each of its programs with warnings of potential errors and tips to improve their code, and will offer teachers a tool to assist them in correcting their students proyects.
+In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and by analyzing a much larger number of projects to achieve more accurate conclusions.
+In this sense, the release of a research dataset with the first five years of data from the Scratch repository\footnote{\url{https://sites.google.com/site/scratchdatameeting/}} will be certainly a great help to our work.
+Finally, we are in the process of developing a website that will allow students to receive immediate feedback for their programs with warnings of potential errors and tips to improve their code.
 
@@
 been funded in part by the Spanish Government under project 
-SobreSale (TIN2011-28110) and by the Comunidad de Madrid 
-under project eMadrid (FIXME).
+SobreSale (TIN2011-28110) and under project ``eMadrid - Investigación y Desarrollo de tecnologías para el e-learning en la Comunidad de Madrid'' (S2009/TIC-1650) funded by the Region of Madrid.
 

### 2014-04-23 paperFIE_en.tex
Add results table. Add comment regarding length of programs to be considered equal.

@@
 In our work as instructors with high school students, we have detected these and other bad practices, such as the repetition of code and object naming, on a regular basis.
-This paper focuses on verifying whether these issues can be generally be found in the Scratch community, by analyzing a large number of projects available on the Scractch community website.
+This paper focuses on verifying whether these issues can generally be found in the Scratch community, by analyzing a large number of projects available on the Scractch community website.
 
@@
   \item \textit{duplicate.DuplicateScripts}\footnote{\url{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py}} analyzes a Scratch project to try to locate repeated complete programs within a project.
-In order to do so, the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal.
+In order to do so, the plug-in runs each script of the project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal.
+A minimum length of five blocks is applied for programs to be considered duplicate rather than coincidentally similar.
+
   To illustrate the operation of this plug-in, Fig.~\ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.
@@
 
-The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects.
+The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects, as shown in table \ref{table:results}.
 
@@
 FIXME: here comes the table with some more results. Important: to note how the distribution (talk about mean and median).
+
+\begin{table}
+\begin{center}
+  \begin{tabular}{ | p{2.5cm} | p{1cm} | p{1.2cm} | p{1cm} |}
+   \hline
+     & Default names & Duplicated scripts & Defined blocks \\ \hline\hline
+    Projects incurring in & 79 & 62 & 17 \\ \hline
+    Mean & 5.94 & 7.23 & 1.11 \\ \hline
+    Median & 3 & 2 & 0 \\ \hline
+    Maximum & 67 & 71 & 25 \\
+    \hline    
+  \end{tabular}
+\end{center}
+\caption{Table of results}
+\label{table:results}
+\end{table}
 

### 2014-04-23 paperFIE_en.tex
commenting the results. balancing the columns in the final page. passing aspell. It is now a release candidate

@@
 \usepackage{url}
+\clubpenalty = 10000
+\widowpenalty = 10000
+\displaywidowpenalty = 10000
+\usepackage{flushend}
 
@@
 In our work as instructors with high school students, we have detected these and other bad practices, such as the repetition of code and object naming, on a regular basis.
-This paper focuses on verifying whether these issues can generally be found in the Scratch community, by analyzing a large number of projects available on the Scractch community website.
+This paper focuses on verifying whether these issues can generally be found in the Scratch community, by analyzing a large number of projects available on the Scratch community website.
 
@@
 As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created for this version.
-A replication package of the projects under study, and the specific version of the plug-ins used in it can be obtained publicly\footnote{FIXME: URL \url{http://gsyc.urjc.es/~grex/repro/2014-fie-scratch}}.
+A replication package of the projects under study, and the specific version of the plug-ins used in it can be obtained publicly\footnote{\url{http://gsyc.urjc.es/~grex/repro/2014-fie-scratch}}.
 
@@
 
-The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects, as shown in table \ref{table:results}.
-
-Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, 79\% of the analyzed projects contained at least one non-renamed \textit{SpriteX} character.
-
-As for the repetition code, 62\% of the analyzed projects contained at least one repeated program.
-In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project.
-
-FIXME: here comes the table with some more results. Important: to note how the distribution (talk about mean and median).
+The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects, as shown in Table~\ref{table:results}.
 
@@
 \begin{center}
-  \begin{tabular}{ | p{2.5cm} | p{1cm} | p{1.2cm} | p{1cm} |}
+  \begin{tabular}{ | l | r | r | r |}
    \hline
-     & Default names & Duplicated scripts & Defined blocks \\ \hline\hline
-    Projects incurring in & 79 & 62 & 17 \\ \hline
+              & Default names & Duplicated scripts & Blocks defined \\ \hline\hline
+    Projects & 79 & 62 & 17 \\ \hline
     Mean & 5.94 & 7.23 & 1.11 \\ \hline
@@
 \end{table}
+
+Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, 79\% of the analyzed projects contained at least one non-renamed \textit{SpriteX} character.
+The maximum number of non-renamed \textit{SpriteX}s found has been 67, which makes the code very hard to maintain (i.e., one could consider this almost as \emph{code obfuscation}).
+In mean, almost 6 \textit{SpriteX}s can be found in a project, while the median is 3.
+Our results show that this is a severe problem in the analyzed projects in Scratch, as in addition to the fact that re-naming occurs in very few projects, the number of elements used in the others are significant in number; being those programs are very hard to understand, its maintenance and reuse are hindered.
+
+As for the repetition code, 62\% of the analyzed projects contained at least one repeated program.
+In fact, only 17\% of the analyzed projects made use of block \textit{def\_block} to define its own blocks that can be reused in other parts of the project.
+The mean of duplicated scripts and blocks defined provides us insight about how difficult is for the authors of the analyzed projects to use abstraction and modularization: duplicated code appears over six times more than blocks.
 

### 2014-10-18 slides/slides.tex
Add slides folder

%
% $Id: $
%
%
% Compilar a .pdf con LaTeX (pdflatex)
% Es necesario instalar Beamer (paquete latex-beamer en Debian)
%

%
% Gráficos:
% Los gráficos pueden suministrarse en PNG, JPG, TIF, PDF, MPS
% Los EPS deben convertirse a PDF (usar epstopdf)
%

\documentclass{beamer}
\usetheme{Warsaw}
%\usebackgroundtemplate{\includegraphics[width=\paperwidth]{format/libresoft-bg.png}}
%\usepackage[spanish]{babel}
\usepackage[latin1]{inputenc}
\usepackage{graphics}
\usepackage{amssymb} % Simbolos matematicos
\usepackage{url}

%\definecolor{libresoftgreen}{RGB}{162,190,43}
%\definecolor{libresoftblue}{RGB}{0,98,143}

%\setbeamercolor{titlelike}{bg=libresoftgreen}

%% Metadatos del PDF.
\hypersetup{
  pdftitle={Automatic Detection of Bad Programming Habits in Scratch: A Preliminary Study},
  pdfauthor={Jesús Moreno León, Gregorio Robles},
  pdfcreator={GSyC/LibreSoft \\ Universidad Rey Juan Carlos},
  pdfproducer=PDFLaTeX,
  pdfsubject={Learning to code with Scratch. Automatic assesment.},
}
%%

\begin{document}

\title{Automatic Detection of Bad Programming Habits in Scratch}
\subtitle{A Preliminary Study}
\institute{jesus.moreno@programamos.es, grex@gsyc.urjc.es \\
GSyC/Libresoft, Universidad Rey Juan Carlos}
\author{Jesús Moreno León, Gregorio Robles}
\date{FIE 2014, Madrid, October 23 2014}

\frame{
\maketitle
\begin{center}
\includegraphics[width=2cm]{format/libresoft-logo}
\hspace{0.5cm}
\includegraphics[width=5cm]{format/gsyc-urjc}
\vspace{0.5cm}
\includegraphics[width=3cm]{format/emadrid.png}
\end{center}
}

% Si el titulo o el autor se quieren acortar para los pies de página
% se pueden redefinir aquí:
%\title{Titulo corto}
%\author{Autores abreviado}

%% LICENCIA DE REDISTRIBUCION DE LAS TRANSPAS
\frame{
~
\vspace{3cm}

\begin{flushright}
\includegraphics[width=2.2cm]{figs/by-sa}

{\tiny
(cc) 2014 Gregorio Robles and Jesús Moreno León\\
  Some rights reserved. This work licensed under Creative Commons \\
  Attribution-ShareAlike License. To view a copy of full license, see \\
  http://creativecommons.org/licenses/by-sa/3.0/ or write to \\
  Creative Commons, 559 Nathan Abbott Way, Stanford, \\
  California 94305, USA. \\
\ \\
Some of the figures have been taken from the Internet \\
Source, and author and licence if known, is specified. \\
For those images, \emph{fair use} applies.
}
\end{flushright}
}
%%

\section{About}

%--------------------------------------------------------
\usebackgroundtemplate{\includegraphics[width=19cm]{figs/AngryBirds.png}}

\begin{frame}
\frametitle{Goal of our paper}

\begin{center}
\Huge Are bad programming habits a common practice in the Scratch community?
\end{center}

\end{frame}

%-----------------------    ---------------------------------
\usebackgroundtemplate{\includegraphics[width=14cm]{figs/audience.png}}
% http://cdn.netrafic.com/wp-content/uploads/2011/02/audience.gif

\begin{frame}
\frametitle{Audience}

Who should/could be interested in this talk?

\begin{itemize}
  \item Educators teaching to code
  \item Students learning to code
  \item Programming learning tools developers
\end{itemize}

\end{frame}

\usebackgroundtemplate{}
%--------------------------------------------------------
\begin{frame}
\frametitle{Scratch}

  \begin{columns}[T]
    \begin{column}{1\textwidth}
     \begin{block}{Learning to code with Scratch}
\begin{itemize}
  \item Success to teach both basic and advanced programming concepts
  \item Bad programming habits detected
  \item Lack of automatic tools to check for \textit{correctness}
  \item Hairball: Lint-inspired Static Analysis of Scratch Projects
\end{itemize}
    \end{block}
    \end{column}

  \end{columns}

\end{frame}

\usebackgroundtemplate{}

%--------------------------------------------------------
\begin{frame}
\frametitle{Bad programming habits with Scratch: objects naming}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=7cm, height=6cm]{figs/SpriteNaming.png}
\end{center}
\label{fig:naming}
\end{figure}

\begin{center}
Sprites naming
\end{center}
\end{frame}

%--------------------------------------------------------
\begin{frame}
\frametitle{Bad programming habits with Scratch: repetition code}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=6cm,height=2.7cm]{figs/CodeRepetition1.png}
\end{center}
\label{fig:repetition1}
\end{figure}

\begin{center}
Repetition code
\end{center}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=6cm,height=2.7cm]{figs/CodeRepetition2.png}
\end{center}
\label{fig:repetition2}
\end{figure}

\begin{center}
Creating blocks to avoid repetition code
\end{center}

\end{frame}

%--------------------------------------------------------
%--------------------------------------------------------
\usebackgroundtemplate{\includegraphics[width=13cm,height=9.2cm]{figs/plugins.png}}
% background: http://25.media.tumblr.com/b83aa72682992ab34b8ce7e61c0cb7f9/tumblr_menxc7qcq61ryin08o1_r1_1280.jpg
\begin{frame}
\frametitle{Hairball plug-ins development}

We developed two plug-ins for Hairball to automatically detect bad programming habits:
\begin{itemize}
  \item convention.SpriteNaming
  \item duplicate.DuplicateScripts
\end{itemize}

\end{frame}

\usebackgroundtemplate{}

%--------------------------------------------------------
%\usebackgroundtemplate{\includegraphics[width=13cm]{figs/iceberg.jpg}}
% background: http://www.wim-network.org/wp-content/uploads/2012/04/iceberg.jpg

\begin{frame}
\frametitle{Scratch projects repository analysis}

\begin{table}
\begin{center}
  \begin{tabular}{ | l | r | r | r |}
   \hline
              & Default names & Duplicated scripts & Defined blocks \\ \hline\hline
    Projects & 79 & 62 & 17 \\ \hline
    Mean & 5.94 & 7.23 & 1.11 \\ \hline
    Median & 3 & 2 & 0 \\ \hline
    Maximum & 67 & 71 & 25 \\
    \hline    
  \end{tabular}
\end{center}
\caption{Analysis of 100 ramdonly downloaded projects}
\label{table:results}
\end{table}
\end{frame}

%--------------------------------------------------------
%\usebackgroundtemplate{\includegraphics[width=13cm]{figs/take-away.jpg}}
% background: http://2.bp.blogspot.com/-78Eh4TBpdtU/UPw7ULV73PI/AAAAAAAAHAE/6DQfvPNCo-Y/s1600/8723052-stylized-red-stamp-showing-the-term-take-away-all-on-white-background.jpg
\usebackgroundtemplate{\includegraphics[width=13cm]{figs/future.png}}

\begin{frame}
\frametitle{Future Work}

\begin{itemize}
  \item Extend the scope of the study developing new plug-ins
  \item Analyze dataset with 5 years data from Scratch website
  \item Dr.Scratch: http://drscratch.programamos.es
\end{itemize}
\vspace{\baselineskip}
\vspace{\baselineskip}
\hfill{\Tiny Background picture: Simon Cunningham }

\end{frame}

%--------------------------------------------------------
%\begin{frame}
%\frametitle{GSyC/LibreSoft}

%\begin{figure}[t!]
%\begin{center}
%\includegraphics[width=11cm]{figs/libresoft.jpg}
%\end{center}
%\label{fig:libresoft}
%\end{figure}

%\begin{center}
%The GSyC/LibreSoft research team at URJC (Madrid).
%\end{center}

%\end{frame}

%--------------------------------------------------------
%\begin{frame}
%\frametitle{Our work at GSyC/LibreSoft}

%\begin{figure}[t!]
%\begin{center}
%\includegraphics[width=2.5cm]{figs/research}
% http://www.memphis.edu/crow/images/research_2.jpg
%\hspace{0.1cm}
%\includegraphics[width=2.5cm]{figs/teaching}
% http://2.bp.blogspot.com/_uxgwfriLwSo/TOWDr8IjaLI/AAAAAAAABLM/d7H-G5jIq-c/s1600/teaching.gif
%\hspace{0.1cm}
%\includegraphics[width=2.5cm]{figs/development}
% http://www.vidadigitalradio.com/wp-content/uploads/2009/04/hackers_cartoons.jpg
%\hspace{0.1cm}
%\includegraphics[width=2.5cm]{figs/promotion}
% http://bloggeate.com/wp-content/uploads/2011/04/como-promocionar-tu-blog.jpg
%\end{center}
%\label{fig:whatwedo}
%\end{figure}

%\begin{center}
%GSyC/LibreSoft's tasks: research, teaching, development, promotion of free software.

%\end{center}
%\end{frame}

\usebackgroundtemplate{}

%--------------------------------------------------------
\frame{
\maketitle
\begin{center}
\includegraphics[width=2cm]{format/libresoft-logo}
\hspace{0.5cm}
\includegraphics[width=5cm]{format/gsyc-urjc}
\vspace{0.5cm}
\includegraphics[width=3cm]{format/emadrid.png}
\end{center}
}

\end{document}

### 2014-10-18 slides/slides.tex
Some changes here and there

@@
 
-\section{About}
+\section{FIE 2014 - eMadrid Session}
 
@@
 \begin{center}
-\Huge Are bad programming habits a common practice in the Scratch community?
+\Huge {\bf Are bad programming habits a common practice in the Scratch community?}
 \end{center}
@@
 \begin{itemize}
-  \item Educators teaching to code
-  \item Students learning to code
-  \item Programming learning tools developers
+  \item Educators teaching how to code
+  \item Students learning to program
+  \item Developers of programming learning tools
 \end{itemize}
@@
 \begin{itemize}
-  \item Success to teach both basic and advanced programming concepts
-  \item Bad programming habits detected
-  \item Lack of automatic tools to check for \textit{correctness}
-  \item Hairball: Lint-inspired Static Analysis of Scratch Projects
+  \item Scratch has shown to be successfull in teaching basic and advanced programming concepts
+  \item However, bad programming habits have been detected
+  \item There are no automatic tools to check for \textit{correctness}
+  \item \texttt{Hairball}: \texttt{lint}-inspired static analysis of Scratch projects
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Bad programming habits with Scratch: objects naming}
+\frametitle{Bad programming habits with Scratch (I)}
 
@@
 \begin{center}
-Sprites naming
+\emph{Bad}/default naming of sprites
 \end{center}
@@
 \begin{frame}
-\frametitle{Bad programming habits with Scratch: repetition code}
+\frametitle{Bad programming habits in Scratch (and II)}
 
+  \begin{columns}[T]
+    \begin{column}{0.5\textwidth}
+
+     \begin{block}{Example of repeated code}
 \begin{figure}[t!]
 \begin{center}
-\includegraphics[width=6cm,height=2.7cm]{figs/CodeRepetition1.png}
+\includegraphics[width=5.4cm,height=2.5cm]{figs/CodeRepetition1.png}
 \end{center}
@@
 \end{figure}
-
-\begin{center}
-Repetition code
-\end{center}
-
+     \end{block}
+    \end{column}
+    \begin{column}{0.5\textwidth}
+     \begin{block}{Solution to avoid repeated code}
 \begin{figure}[t!]
 \begin{center}
-\includegraphics[width=6cm,height=2.7cm]{figs/CodeRepetition2.png}
+\includegraphics[width=5.4cm,height=2.5cm]{figs/CodeRepetition2.png}
 \end{center}
+Blocks should be created to avoid repetition of code
 \label{fig:repetition2}
 \end{figure}
+     \end{block}
 
-\begin{center}
-Creating blocks to avoid repetition code
-\end{center}
+    \end{column}
+  \end{columns}
 
@@
 
-We developed two plug-ins for Hairball to automatically detect bad programming habits:
-\begin{itemize}
+  \begin{columns}[T]
+    \begin{column}{0.8\textwidth}
+     \begin{block}{\vspace{0.1cm}\center{We have developed two plug-ins for Hairball to automatically detect bad programming habits}\vspace{0.1cm}}
+\vspace{0.3cm}
+\begin{enumerate}
   \item convention.SpriteNaming
+\vspace{0.15cm}
   \item duplicate.DuplicateScripts
-\end{itemize}
+\end{enumerate}
+\vspace{0.4cm}
+    \end{block}
+    \end{column}
+  \end{columns}
 
@@
 \begin{center}
-  \begin{tabular}{ | l | r | r | r |}
+  \begin{tabular}{ | l | c | c | c |}
    \hline
@@
     Mean & 5.94 & 7.23 & 1.11 \\ \hline
-    Median & 3 & 2 & 0 \\ \hline
+ 22   Median & 3 & 2 & 0 \\ \hline
     Maximum & 67 & 71 & 25 \\
@@
 \end{center}
-\caption{Analysis of 100 ramdonly downloaded projects}
+\caption{Analysis of 100 ramdonly downloaded Scratch projects}
 \label{table:results}
@@
 
-\begin{itemize}
+\begin{enumerate}
   \item Extend the scope of the study developing new plug-ins
-  \item Analyze dataset with 5 years data from Scratch website
-  \item Dr.Scratch: http://drscratch.programamos.es
-\end{itemize}
+  \item Analyze dataset with 5 years of data from the Scratch website
+  \item Dr. Scratch (alpha version): http://drscratch.programamos.es
+\end{enumerate}
 \vspace{\baselineskip}

### 2014-10-20 slides/slides.tex
Fix errata (median)

@@
     Mean & 5.94 & 7.23 & 1.11 \\ \hline
- 22   Median & 3 & 2 & 0 \\ \hline
+    Median & 3 & 2 & 0 \\ \hline
     Maximum & 67 & 71 & 25 \\
